---
title: "Building the analysis dataset"
subtitle: "From the raw extraction CSVs to one row per activity"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: show
execute:
  warning: false
  message: false
---

Everything here reads CSVs committed to `data/interim/`, written once by
extraction/01_frame. Nothing in this document touches `iati.sqlite` — that
is the whole point of the split: this script runs to completion on a clean
clone of the repository, with no access to the 32 GB source.

```{r}
library(tidyverse)

read_gbs <- function(name, ...) read_csv(file.path("../data/interim", name), ...)
```

## 0. Analytical decisions

Two restrictions live here rather than in extraction, because both are
analytical choices that may change without the source data changing. Editing
either is a one-line change plus a re-render; encoding them upstream would
require re-downloading 32 GB to revisit.

```{r}
# 2014 is where IATI reporting becomes regular enough to treat a missing
# disbursement as informative rather than as an artefact of non-reporting.
# The raw transaction extract reaches back to 1989 for some activities.
YEAR_MIN <- 2014

# The analytical window closes in 2023 because the World Bank publishes net ODA
# as a share of GNI with a two-year lag, and that is the aid-dependence control
# the suspension literature identifies as predictive. Raise this when
# the 2024 figures are released.
YEAR_MAX <- 2023

# Drop activities whose parent is also in the frame: the same programme money
# is otherwise counted at both programme and contract level, inflating both the
# numerator and the denominator of the donor-share variable. Set to
# FALSE to test how much the exclusion matters.
DROP_CHILD_ACTIVITIES <- TRUE
```

## 1. Frame and donor type

```{r}
frame_raw <- read_gbs("gbs_frame.csv", col_types = cols(
  iatiidentifier = col_character(),
  reportingorg_ref = col_character(),
  reportingorg_type = col_character(),
  reportingorg_narrative = col_character(),
  activitystatus_code = col_character(),
  has_a01 = col_logical(),
  has_a02 = col_logical(),
  declared_activity_level = col_logical(),
  declared_transaction_level = col_logical(),
  .default = col_guess()
))

# org_type 15 ("other public sector") sits with the multilaterals: it covers
# entities like the Green Climate Fund, not bilateral government agencies
#.
x1_lookup <- c("10" = "Bilateral", "15" = "Multilateral", "40" = "Multilateral")

status_lookup <- c(
  "1" = "Pipeline/identification", "2" = "Implementation",
  "3" = "Finalisation", "4" = "Closed", "5" = "Cancelled"
)

frame <- frame_raw |>
  transmute(
    iatiidentifier,
    donor_ref = reportingorg_ref,
    donor     = reportingorg_narrative,
    org_type  = reportingorg_type,
    x1_donor_type = x1_lookup[org_type],
    a01 = as.integer(has_a01), a02 = as.integer(has_a02),
    # A02-only is "sector budget support"; A01-only "general"; both is rare
    # enough (17 activities corpus-wide) to keep as its own label rather than
    # force a choice.
    modality = case_when(
      a01 == 1 & a02 == 1 ~ "both",
      a02 == 1             ~ "A02_sector",
      a01 == 1             ~ "A01_general"
    ),
    declared_activity    = as.integer(declared_activity_level),
    declared_transaction = as.integer(declared_transaction_level),
    status_code = activitystatus_code,
    status_name = status_lookup[status_code]
  )

frame |> count(x1_donor_type, modality)
```

Applying the parent-child exclusion declared in §0. Reported as a count rather
than applied silently, so the cost of the decision is visible in the rendered
document.

```{r}
related <- read_gbs("gbs_related.csv", col_types = cols(
  child_id = col_character(), parent_id = col_character()
))

if (DROP_CHILD_ACTIVITIES) {
  n_before <- nrow(frame)
  frame <- frame |> filter(!iatiidentifier %in% related$child_id)
  tibble(
    activities_before = n_before,
    children_dropped  = n_before - nrow(frame),
    activities_after  = nrow(frame)
  )
}
```

## 2. Main recipient country

An activity can list several recipient countries with a percentage each. The
main recipient is the one with the largest share; this affects 54 of the
frame's activities.

```{r}
recipients <- read_gbs("gbs_recipients.csv", col_types = cols(
  iatiidentifier = col_character(), country = col_character(),
  percentage = col_double()
))

recipient_main <- recipients |>
  # A missing percentage cannot win a max() comparison; treating it as 0
  # rather than dropping the row keeps single-country activities in scope
  # even when the percentage element was left blank.
  mutate(percentage = replace_na(percentage, 0)) |>
  slice_max(percentage, n = 1, by = iatiidentifier, with_ties = FALSE) |>
  select(iatiidentifier, country)
```

## 3. Dates, and the outcome they support directly

```{r}
dates <- read_gbs("gbs_dates.csv", col_types = cols(
  iatiidentifier = col_character(), date_type = col_character(),
  isodate = col_character()
)) |>
  filter(!is.na(date_type)) |>
  distinct(iatiidentifier, date_type, .keep_all = TRUE) |>
  pivot_wider(names_from = date_type, values_from = isodate)

# Early closure: the convergent-validity check for the financial gap.
# A positive value means the operation concluded before its announced date.
dates <- dates |>
  mutate(
    across(c(planned_start, actual_start, planned_end, actual_end), as.Date),
    y2_early_days = as.integer(planned_end - actual_end),
    planned_duration_days = as.integer(planned_end - planned_start)
  )
```

## 4. Budget

```{r}
budget_tot <- read_gbs("gbs_budget.csv", col_types = cols(
  iatiidentifier = col_character(), budget_orig = col_double(),
  n_budget_lines = col_integer(), budget_ccy = col_character()
))
```

## 5. Transactions: windowed and full-life

`gbs_tx.csv` holds every commitment (type 2) and disbursement (type 3)
transaction belonging to a framed activity, with no date restriction — the
earliest reaches back to 1989. The analytical window declared in §0 is applied
here, so the panel and every total derived from it span the same years as the
controls.

`gbs_lifecycle.csv` carries the same two totals **without** any window. The two
therefore diverge wherever an activity has transactions outside 2014-2023, and
that divergence is informative rather than an error: it is the difference
between what the activity did in the observed period and what it did in total.

```{r}
tx <- read_gbs("gbs_tx.csv", col_types = cols(
  iatiidentifier = col_character(), ttype = col_character(),
  year = col_integer(), tdate = col_character(), usd = col_double()
)) |>
  filter(between(year, YEAR_MIN, YEAR_MAX))

panel_year <- tx |>
  summarise(
    commitment_usd   = sum(usd[ttype == "2"], na.rm = TRUE),
    disbursement_usd = sum(usd[ttype == "3"], na.rm = TRUE),
    n_disb_tx        = sum(ttype == "3"),
    .by = c(iatiidentifier, year)
  )

activity_tx <- panel_year |>
  summarise(
    commitment_usd   = sum(commitment_usd),
    disbursement_usd = sum(disbursement_usd),
    n_disb_years      = sum(disbursement_usd > 0),
    # An activity with commitment but no disbursement year has an empty vector
    # here; min()/max() on empty warn and return +-Inf rather than NA.
    first_disb_year   = if (any(disbursement_usd > 0)) min(year[disbursement_usd > 0]) else NA_integer_,
    last_disb_year    = if (any(disbursement_usd > 0)) max(year[disbursement_usd > 0]) else NA_integer_,
    .by = iatiidentifier
  )

lifecycle <- read_gbs("gbs_lifecycle.csv", col_types = cols(
  iatiidentifier = col_character(), commitment_usd_full = col_double(),
  disbursement_usd_full = col_double(), n_disb_years_full = col_integer()
))
```

## 6. Controls

```{r}
# The `_name` columns carry IATI's own label for each code, so the codes stay
# numeric for modelling while the labels make tables readable — the same
# code/name pairing `status_code`/`status_name` already uses.
controls <- read_gbs("gbs_controls.csv", col_types = cols(.default = col_character())) |>
  mutate(across(c(finance_type, flow_type, tied_status, collaboration_type,
                  hierarchy), as.integer)) |>
  # Blank labels are the source's own gaps, not a reading artefact.
  mutate(across(ends_with("_name"), ~ na_if(.x, "")))

# The `_name` columns arrive only from a re-rendered 01_frame; guard so an
# older extract does not break the build.
if ("finance_type_name" %in% names(controls)) {
  controls |> count(finance_type, finance_type_name, sort = TRUE)
} else {
  controls |> count(finance_type, sort = TRUE)
}
```

## 7. Assemble `analysis`

One row per activity. `complete_flag` restricts the gap to activities where it
is interpretable — an activity still under implementation can show a large gap
for no reason other than that its disbursements have not happened yet.

```{r}
analysis <- frame |>
  left_join(recipient_main |> rename(recipient_main = country), by = "iatiidentifier") |>
  left_join(activity_tx,  by = "iatiidentifier") |>
  left_join(dates,        by = "iatiidentifier") |>
  left_join(budget_tot,   by = "iatiidentifier") |>
  left_join(lifecycle,    by = "iatiidentifier") |>
  left_join(controls,     by = "iatiidentifier") |>
  mutate(
    cp_eligible   = as.integer(modality == "A01_general"),
    complete_flag = as.integer(status_code %in% c("3", "4")),
    y1_gap        = if_else(commitment_usd > 0, 1 - disbursement_usd / commitment_usd, NA_real_),
    y1_gap_full   = if_else(commitment_usd_full > 0,
                             1 - disbursement_usd_full / commitment_usd_full, NA_real_)
  )

nrow(analysis)
analysis |> filter(complete_flag == 1, !is.na(y1_gap_full)) |>
  summarise(n = n(), n_donors = n_distinct(donor_ref))
```

## 8. Save

```{r}
write_csv(analysis,       "../data/interim/analysis.csv")
write_csv(panel_year,     "../data/interim/panel_year.csv")
write_csv(recipient_main, "../data/interim/recipient_main.csv")
```
