---
title: "External controls for the recipient country-year"
subtitle: "Governance, aid dependence and economic size from the World Bank API"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: show
execute:
  warning: false
  message: false
---

Retrieves the recipient-side controls the design requires, at country-year level,
directly from the World Bank API.

**Why these controls are not optional.** Molenaers, Gagiano, Smets and Dellepiane
(2015) find budget support suspensions track deteriorating voice-and-accountability
trends and lower aid dependence. Without those variables in the model, any
association between a donor's structural position and withholding is open to the
objection that it reflects political rupture in the recipient rather than anything
about the donor. This is the rival explanation the literature review commits to
addressing, so the controls are part of the design rather than a robustness check.

**Why the API rather than a packaged extract.** The indicator codes, their coverage
and their revision dates are all recoverable from the request itself, which makes
the extraction auditable: a reader can re-run this document and obtain the same
table, or see precisely where the source has changed. Downloading a spreadsheet
by hand would lose that.

```{r}
#| label: setup
library(tidyverse)
library(httr2)

# The frame runs 2014 to June 2026.
YEAR_MIN <- 2014
YEAR_MAX <- 2026

dir_external <- "../data/external"
```

## 1. Country crosswalk

The IATI frame identifies recipients by ISO-2 code; the World Bank returns ISO-3.
The crosswalk is taken from the API's own country endpoint rather than a static
lookup, so that it carries the same revision date as the indicator data.

The endpoint returns 295 entities, of which roughly a quarter are aggregates
(regions, income groups). These are dropped: an aggregate cannot be a budget
support recipient and would silently duplicate country-years if it survived the
join.

```{r}
#| label: countries
path_countries <- file.path(dir_external, "wb_countries.csv")

if (file.exists(path_countries)) {
  countries <- read_csv(path_countries, col_types = cols(
    iso3         = col_character(),
    iso2         = col_character(),
    country_name = col_character(),
    region       = col_character(),
    income_group = col_character()
  ))
} else {
  resp <- request("https://api.worldbank.org/v2/country") |>
    req_url_query(format = "json", per_page = 400) |>
    req_retry(max_tries = 3) |>
    req_perform() |>
    resp_body_json(simplifyVector = FALSE)

  countries <- resp[[2]] |>
    map(\(x) tibble(
      iso3         = x$id,
      iso2         = x$iso2Code,
      country_name = x$name,
      region       = x$region$value,
      income_group = x$incomeLevel$value
    )) |>
    list_rbind() |>
    # "Aggregates" is the region label the API gives to every non-country entity.
    filter(region != "Aggregates")

  write_csv(countries, path_countries)
}

countries |> count(region, name = "n_countries")
```

## 2. A single fetch function

One function covers both sources. The Worldwide Governance Indicators moved to
their own source (`source = 3`) and carry a `GOV_WGI_` prefix; the older codes
return "indicator not found", which is a silent failure mode worth naming, since
the API answers with HTTP 200 and an error object in the body rather than a
non-success status.

```{r}
#| label: fetch-fn
fetch_wb_indicator <- function(indicator, source = NULL,
                               year_min = YEAR_MIN, year_max = YEAR_MAX) {

  req <- request("https://api.worldbank.org/v2/country/all/indicator") |>
    req_url_path_append(indicator) |>
    req_url_query(
      format   = "json",
      date     = str_c(year_min, ":", year_max),
      per_page = 20000
    ) |>
    req_throttle(rate = 5) |>
    req_retry(max_tries = 3)

  # The WGI live in a separate source; omitting it returns an empty result
  # rather than an error.
  if (!is.null(source)) req <- req_url_query(req, source = source)

  body <- req |> req_perform() |> resp_body_json(simplifyVector = FALSE)

  # The API signals a bad indicator code inside a 200 response, so the shape of
  # the body is the only way to detect it.
  if (is.null(body[[1]]$total)) {
    stop("World Bank API returned no metadata for '", indicator,
         "'. Most likely the code is retired.", call. = FALSE)
  }

  records <- body[[2]]
  if (is.null(records) || length(records) == 0) {
    return(tibble(iso3 = character(), year = integer(), value = double()))
  }

  records |>
    map(\(x) tibble(
      iso3  = x$countryiso3code %||% NA_character_,
      year  = as.integer(x$date),
      value = if (is.null(x$value)) NA_real_ else as.numeric(x$value)
    )) |>
    list_rbind() |>
    filter(!is.na(iso3), iso3 != "")
}
```

## 3. The indicator set

Each indicator earns its place from a specific claim in the literature rather
than from availability.

```{r}
#| label: indicator-table
indicators <- tribble(
  ~name,              ~code,                    ~source, ~rationale,
  "voice_account",    "GOV_WGI_VA.EST",         "3",     "Molenaers et al. (2015): suspensions track deteriorating voice and accountability",
  "corruption_ctrl",  "GOV_WGI_CC.EST",         "3",     "Corruption is the single largest stated breach category in the suspension dataset",
  "pol_stability",    "GOV_WGI_PV.EST",         "3",     "Proxies the political rupture that is the main rival explanation",
  "aid_dependence",   "DT.ODA.ODAT.GN.ZS",      NA,      "Molenaers et al. (2015): aid dependence reduces suspension likelihood",
  "gdp_pc",           "NY.GDP.PCAP.CD",         NA,      "Recipient income level; conventional control",
  "population",       "SP.POP.TOTL",            NA,      "Scale; enters logged"
)

indicators |> select(name, code, rationale)
```

## 4. Retrieval

Cached to disk on first run. The API is not re-queried on subsequent renders,
which keeps the document reproducible and avoids repeated load on a public
service.

```{r}
#| label: fetch
path_controls <- file.path(dir_external, "wb_controls.csv")

if (file.exists(path_controls)) {
  controls_long <- read_csv(path_controls, col_types = cols(
    name  = col_character(),
    iso3  = col_character(),
    year  = col_integer(),
    value = col_double()
  ))
} else {
  controls_long <- indicators |>
    select(name, code, source) |>
    pmap(\(name, code, source) {
      fetch_wb_indicator(code, source = if (is.na(source)) NULL else source) |>
        mutate(name = name, .before = 1)
    }) |>
    list_rbind()

  write_csv(controls_long, path_controls)
}

controls_long |>
  group_by(name) |>
  summarise(
    country_years = n(),
    non_missing   = sum(!is.na(value)),
    first_year    = min(year),
    last_year     = max(year),
    .groups = "drop"
  )
```

## 5. Coverage, and the gap the design has to absorb

The governance indicators are published with a lag. The frame runs to June 2026;
the WGI series ends earlier. This is reported here rather than patched, because
the choice of how to handle it — restrict the analytical window, carry the last
observation forward, or treat late years as missing — belongs to the modelling
stage and should be visible when it is made.

```{r}
#| label: coverage-by-year
controls_long |>
  filter(!is.na(value)) |>
  count(name, year) |>
  pivot_wider(names_from = name, values_from = n, values_fill = 0) |>
  arrange(year)
```

```{r}
#| label: fig-coverage
#| fig-cap: "Countries with a non-missing observation, by indicator and year"
controls_long |>
  filter(!is.na(value)) |>
  count(name, year) |>
  ggplot(aes(year, n, colour = name)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.2) +
  scale_x_continuous(breaks = seq(YEAR_MIN, YEAR_MAX, 2)) +
  labs(x = NULL, y = "Countries reporting", colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## 6. Wide form, keyed for the join

Written on ISO-2 as well as ISO-3, since the IATI frame keys on ISO-2 and the
join should not require the crosswalk to be repeated downstream.

```{r}
#| label: reshape
controls <- controls_long |>
  pivot_wider(names_from = name, values_from = value) |>
  left_join(countries |> select(iso3, iso2, country_name, region, income_group),
            by = "iso3") |>
  # A country in the indicator series but absent from the country endpoint is
  # an aggregate that survived; dropping it here keeps the join one-to-one.
  filter(!is.na(iso2)) |>
  relocate(iso2, iso3, country_name, year)

controls |> glimpse()
```

```{r}
#| label: save
write_csv(controls, file.path(dir_external, "controls_country_year.csv"))
```

**Output.** `data/external/controls_country_year.csv`, keyed on `iso2` × `year`,
ready to join to the activity-year panel built in 01_build_tables.
