---
title: "Political-rupture controls"
subtitle: "Coups and election timing for the recipient country-year"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: show
execute:
  warning: false
  message: false
---

Two controls for recipient-side political rupture, both properties of the
recipient country alone rather than of the donor-recipient pair. This is what
makes them usable across the whole sample: donor-recipient ideological
alignment was considered and rejected for exactly the opposite reason — it is
undefined for 91 per cent of the analytical sample, whose donor is a
multilateral institution rather than a state.

Source: [V-Dem](https://v-dem.net) via the `vdemdata` package, installed from
GitHub rather than CRAN. It is used because no current, freely downloadable,
registration-free source of coup and election-year data exists elsewhere.

```{r}
#| label: setup
library(tidyverse)

if (!requireNamespace("vdemdata", quietly = TRUE)) {
  remotes::install_github("vdeminstitute/vdemdata")
}
library(vdemdata)

dir_external <- "../data/external"
YEAR_MIN <- 2014
YEAR_MAX <- 2026
```

## 1. Country crosswalk

Same cache 02_external_controls writes — reused if present, built here
if not, so this document does not depend on the other having run first.

```{r}
path_countries <- file.path(dir_external, "wb_countries.csv")

if (file.exists(path_countries)) {
  countries <- read_csv(path_countries, col_types = cols(
    iso3 = col_character(), iso2 = col_character(),
    country_name = col_character(), region = col_character(),
    income_group = col_character()
  ))
} else {
  library(httr2)
  resp <- request("https://api.worldbank.org/v2/country") |>
    req_url_query(format = "json", per_page = 400) |>
    req_retry(max_tries = 3) |>
    req_perform() |>
    resp_body_json(simplifyVector = FALSE)

  countries <- resp[[2]] |>
    map(\(x) tibble(
      iso3 = x$id, iso2 = x$iso2Code, country_name = x$name,
      region = x$region$value, income_group = x$incomeLevel$value
    )) |>
    list_rbind() |>
    filter(region != "Aggregates")

  write_csv(countries, path_countries)
}
```

## 2. Coups

`e_pt_coup` and `e_pt_coup_attempts` are the Powell and Thyne indicators,
bundled into V-Dem and updated on its release schedule — current through 2025,
which is more recent than either standalone mirror located during the search
for this control.

```{r}
#| label: coups
coups <- vdem |>
  filter(year >= YEAR_MIN, year <= YEAR_MAX) |>
  transmute(
    iso3 = country_text_id,
    year,
    coup          = replace_na(as.integer(e_pt_coup > 0), 0L),
    coup_attempt  = replace_na(as.integer(e_pt_coup_attempts > 0), 0L)
  )

coups |> filter(coup == 1) |> count(year, name = "countries_with_coup")
```

## 3. Elections

An election of any type in a given country-year leaves at least one of
`v2eltype_0` through `v2eltype_9` non-missing; which specific slot is filled
depends on how many election types occurred that year; the type codes
themselves are not used here, only whether an election happened at all.

```{r}
#| label: elections
eltype_cols <- paste0("v2eltype_", 0:9)

elections <- vdem |>
  filter(year >= YEAR_MIN, year <= YEAR_MAX) |>
  mutate(election_year = as.integer(if_any(all_of(eltype_cols), ~ !is.na(.x)))) |>
  transmute(iso3 = country_text_id, year, election_year)

elections |> count(election_year)
```

## 4. Coverage against the analytical frame

```{r}
#| label: coverage
political <- coups |>
  full_join(elections, by = c("iso3", "year"))

political |>
  group_by(year) |>
  summarise(
    n_countries = n(),
    pct_coup    = round(100 * mean(coup, na.rm = TRUE)),
    pct_election = round(100 * mean(election_year, na.rm = TRUE)),
    .groups = "drop"
  )
```

The source runs through 2025; H1 2026 activities in the frame carry no
observation for that half-year and are left `NA` rather than backfilled —
carrying forward an unobserved political event would be worse than admitting
it is unknown.

## 5. Attach ISO2 and save

```{r}
#| label: save
political <- political |>
  left_join(countries |> select(iso3, iso2), by = "iso3") |>
  filter(!is.na(iso2)) |>
  relocate(iso2, iso3, year)

write_csv(political, file.path(dir_external, "controls_political.csv"))
political |> glimpse()
```

**Output.** `data/external/controls_political.csv`, keyed on `iso2` x `year`,
joins to the same key as `controls_country_year.csv` from
02_external_controls.
