---
title: "Structural position of donors in recipient budget support"
subtitle: "S1 (share), S2 (fragmentation) and S3 (pool size), built on commitments"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: show
execute:
  warning: false
  message: false
---

Builds the principal explanatory variables for H1, H2 and H3. Reads
`analysis.csv` and `panel_year.csv` from the construction step upstream — no
database connection, so this renders on a clean clone with no access to the
32 GB source.

**The three variables are built on commitments, not disbursements**. Measured
from disbursements, a donor's share is determined by the outcome: a donor that
withholds records a share of zero in that same year, and the regressor would
restate the thing it is meant to explain. Commitments are not reduced by a
decision to withhold — the outstanding commitment is precisely what makes
withholding observable — so the measure is structurally prior to the outcome.

It is also the better construct. The theory defines structural position as how
far a recipient depends on a given donor, which is a question about promised
financing rather than about the timing of cash.

**H2** — withholding becomes more likely as the donor's share rises, because
only then is the loss hard to substitute. **H3** — withholding becomes less
likely as the donor pool fragments, or more likely under the signalling
mechanism; the sign is what adjudicates between them.

```{r}
#| label: setup
library(tidyverse)

analysis   <- read_csv("../data/interim/analysis.csv", show_col_types = FALSE)
panel_year <- read_csv("../data/interim/panel_year.csv", show_col_types = FALSE)
```

## 1. Donor x country x year

An activity may list several recipient countries; 01_build_tables §2
assigns each to its single main recipient, so that is already resolved here.

```{r}
#| label: donor-country-year
dcy <- panel_year |>
  left_join(
    analysis |> select(iatiidentifier, donor_ref, donor, country = recipient_main),
    by = "iatiidentifier"
  ) |>
  filter(!is.na(country)) |>
  summarise(
    # Negative annual sums are refunds and reversals netted within a year. A
    # share cannot rest on a negative numerator, so they are floored rather
    # than dropped — dropping would delete the cell entirely, which is what
    # produced the endogeneity diagnosed earlier.
    commit_year  = sum(pmax(commitment_usd, 0), na.rm = TRUE),
    disb_year    = sum(pmax(disbursement_usd, 0), na.rm = TRUE),
    n_activities = n_distinct(iatiidentifier),
    .by = c(donor_ref, donor, country, year)
  )

dcy
```

## 2. Cumulative commitment — the standing promise

Commitments arrive in lumps: a donor signs for a whole operation once and
disburses over several years. An annual commitment share would therefore read
100 per cent in the signing year and nothing afterwards, describing the
paperwork rather than the relationship. The **cumulative** commitment to date is
what stands in any given year, and that is what "structural" means here.

```{r}
#| label: cumulative
dcy <- dcy |>
  arrange(donor_ref, country, year) |>
  mutate(cum_commit = cumsum(commit_year), .by = c(donor_ref, country))

country_year <- dcy |>
  summarise(
    total_commit = sum(cum_commit),
    n_donors     = n_distinct(donor_ref),
    .by = c(country, year)
  )
```

## 3. S1, S2, S3

```{r}
#| label: structural
structural <- dcy |>
  left_join(country_year, by = c("country", "year")) |>
  mutate(
    # A recipient-year with nothing committed by anyone has no share to
    # compute; that is undefined rather than zero.
    s1_share = if_else(total_commit > 0, cum_commit / total_commit, NA_real_)
  ) |>
  mutate(
    s2_hhi = sum(s1_share^2, na.rm = TRUE),
    # What matters to a donor considering withholding is how substitutable IT
    # is, which depends on the others — hence the pool net of its own weight.
    s2_hhi_net  = if_else(n_donors > 1, s2_hhi - s1_share^2, NA_real_),
    s3_n_donors = n_donors,
    .by = c(country, year)
  ) |>
  select(donor_ref, donor, country, year, commit_year, disb_year, cum_commit,
         total_commit, s1_share, s2_hhi, s2_hhi_net, s3_n_donors, n_activities)

structural
```

## 4. Coverage, and the balance that matters

The diagnostic that decided this specification. If the measure were still
endogenous, it would be missing far more often in years the outcome marks as
withheld than elsewhere.

```{r}
#| label: balance
structural |>
  mutate(no_disbursement = disb_year <= 0) |>
  summarise(
    cells          = n(),
    s1_missing_pct = round(100 * mean(is.na(s1_share))),
    .by = no_disbursement
  )
```

- Missingness is the same on both sides of the outcome. That is the point of
  building on commitments: a year in which the donor sent nothing still has a
  fully observed structural position.

```{r}
#| label: coverage
structural |>
  summarise(
    donor_country_years = n(),
    s1_available        = sum(!is.na(s1_share)),
    s2_net_available    = sum(!is.na(s2_hhi_net)),
    single_donor        = sum(s3_n_donors == 1),
    pct_single          = round(100 * mean(s3_n_donors == 1)),
    median_share        = round(median(s1_share, na.rm = TRUE), 3)
  )
```

- `s2_hhi_net` is undefined wherever the pool holds one donor: a pool net of its
  only member is empty, not zero. Those cells are kept for H2, where a share of
  one is the theoretical extreme of maximal leverage, and dropped from H3, which
  has no fragmentation to measure there.

## 5. Does the key variable vary?

```{r}
#| label: variation
structural |>
  mutate(band = cut(s1_share, c(0, .1, .25, .5, .75, .99, 1),
                    labels = c("<10%", "10-25%", "25-50%", "50-75%", "75-99%", "100%"),
                    include.lowest = TRUE)) |>
  count(band)

structural |> count(s3_n_donors) |> arrange(s3_n_donors)
```

```{r}
#| label: fig-share
#| fig-cap: "Donor share of recipient committed budget support, excluding single-donor cells"
structural |>
  filter(s3_n_donors > 1, !is.na(s1_share)) |>
  ggplot(aes(s1_share)) +
  geom_histogram(bins = 30, fill = "grey30") +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Donor share of recipient committed budget support",
       y = "Donor-country-years") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: fig-commit-vs-disb
#| fig-cap: "Commitment share against realised disbursement share, same cells"
structural |>
  # if_else needs a scalar condition; the group total is one value but the
  # branches are vectors, so the guard goes on the denominator instead.
  mutate(total_disb = sum(disb_year), .by = c(country, year)) |>
  mutate(share_disb = if_else(total_disb > 0, disb_year / total_disb, NA_real_)) |>
  filter(!is.na(s1_share), !is.na(share_disb)) |>
  ggplot(aes(s1_share, share_disb)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linetype = "dashed") +
  geom_point(alpha = 0.25, size = 1) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Share of committed funds", y = "Share of funds actually disbursed") +
  theme_minimal(base_size = 11)
```

- Points below the diagonal are donor-country-years where the donor disbursed a
  smaller share than it had promised — which is what the outcome is built to
  detect. That the two axes diverge is the evidence that they are not the same
  variable, and the reason the commitment side can serve as a predictor of the
  disbursement side.

## 6. Save

```{r}
#| label: save
write_csv(structural, "../data/interim/structural.csv")
```

**Output.** `data/interim/structural.csv`, one row per donor-country-year. The
models use `s1_share`, `s2_hhi_net` and `s3_n_donors`, all built on commitments;
`disb_year` travels with them for the diagnostic above and for the descriptive
tables in analysis/04_eda_clean.
