---
title: "Frame construction"
subtitle: "General budget support operations in the full IATI corpus"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: show
execute:
  warning: false
  message: false
---

Identifies every general budget support operation in the IATI corpus by **aid
modality**, not by purpose code.

The earlier frame filtered on purpose code 51010 and was wrong in both
directions at once. Its undercoverage was invisible from inside the extract: the
World Bank's Development Policy Operations cannot be found by searching a file
that already excluded them. That is why this runs against the full corpus.

Aid type is declared either as an activity-level default or on individual
transactions, and publishers differ in which they use — the AfDB populates only
the transaction level, the EC only the activity level. Taking the union of both
is the difference between a frame and a frame missing half its publishers.

This needs `iati.sqlite` (32 GB), which is not in the repository — see the
README for the download. Every join here runs under ten seconds against a
local copy; the full document renders in one to three minutes.

```{r}
library(tidyverse)
library(DBI)
library(RSQLite)

con <- dbConnect(SQLite(), "C:/iati/iati.sqlite")
```

The source ships with no indexes on the columns this script joins on
repeatedly. Without one, SQLite falls back to a nested-loop scan for the
`recipientcountry` lookup below and the query does not return in reasonable
time; indexed, it runs in about a second. Adding an index only builds a lookup
structure — it does not alter any data, and is safe to leave in place.

```{sql, connection = con}
CREATE INDEX IF NOT EXISTS idx_rc_act ON recipientcountry (_link_activity);
```

## 1. The frame

```{sql, connection = con}
DROP TABLE IF EXISTS gbs_frame;
```

Four restrictions, applied here rather than left implicit: the reporting
organisation must be type 10 (bilateral), 15 (other public sector) or 40
(multilateral) — this excludes NGOs and private-sector reporters that also use
the A01/A02 codes but are not budget support in the theoretical sense; the
activity must have at least one coded recipient country; the US Department of
the Interior (`US-GOV-9`) is excluded by name — its 1,404 A01 activities are
Compacts of Free Association with Micronesia, the Marshall Islands and Palau,
carry no coded recipient country in most cases and are not administered
through a performance assessment framework. Three further reporters are
excluded on the same reasoning: the Netherlands Enterprise Agency
(`NL-KVK-27378529`, 169 activities — sector-specific private-sector trade
programmes, e.g. "Garments and textiles Guatemala", "Rwanda Coffee"); the
Green Climate Fund (`XM-DAC-41317`, discrete project-based climate finance,
e.g. "Bhutan for life", "Jordan Integrated Landscape Management Initiative" —
GCF disburses to accredited implementing entities per project, not to a
recipient treasury under a PAF); and the Netherlands Space Office
(`XI-IATI-NSO`, satellite-data development grants under the G4AW programme,
e.g. "Sat4Business Ghana", "SpiceUp Indonesia" — technical assistance, not
budget support). None of the four excluded reporters engages in policy-based
lending against a negotiated performance assessment framework, which is the
instrument this thesis studies.

```{sql, connection = con}
CREATE TABLE gbs_frame AS
WITH aid_union AS (
    SELECT DISTINCT _link_activity AS act, UPPER(TRIM(code)) AS aid_code,
           'activity' AS declared_at
    FROM defaultaidtype
    WHERE UPPER(TRIM(code)) IN ('A01','A02')
    UNION
    SELECT DISTINCT _link_activity, UPPER(TRIM(code)), 'transaction'
    FROM transaction_aidtype
    WHERE UPPER(TRIM(code)) IN ('A01','A02')
)
SELECT a._link AS act,
       a.iatiidentifier,
       a.reportingorg_ref,
       a.reportingorg_type,
       a.reportingorg_narrative,
       a.activitystatus_code,
       MAX(u.aid_code = 'A01')            AS has_a01,
       MAX(u.aid_code = 'A02')            AS has_a02,
       MAX(u.declared_at = 'activity')    AS declared_activity_level,
       MAX(u.declared_at = 'transaction') AS declared_transaction_level
FROM activity a
JOIN aid_union u ON u.act = a._link
WHERE a.reportingorg_type IN ('10', '15', '40')
  AND a.reportingorg_ref NOT IN
      ('US-GOV-9', 'NL-KVK-27378529', 'XM-DAC-41317', 'XI-IATI-NSO')
  AND EXISTS (SELECT 1 FROM recipientcountry rc WHERE rc._link_activity = a._link)
GROUP BY a._link;
```

16 activities in the corpus carry the **same** `iatiidentifier` under **more
than one** surrogate key (`_link`) — the same activity indexed twice by IATI
Tables, not a genuine duplicate report. Checked directly: every duplicate pair
agrees on donor and activity status, so nothing is lost by keeping one. Left
unresolved, this turns every later join into a many-to-many and silently
inflates row counts — it surfaced exactly that way the first time this ran.

```{sql, connection = con}
DELETE FROM gbs_frame
WHERE act NOT IN (SELECT MIN(act) FROM gbs_frame GROUP BY iatiidentifier);
```

```{sql, connection = con}
CREATE INDEX idx_gbs_frame_act ON gbs_frame (act);
```

A second, larger duplication exists: 452 activities
declare a parent activity — `related-activity` type 1 — that is **also in the
frame**, so the same programme money appears at both the programme and the
contract level. Verified rather than assumed: one EC parent commits USD 68.4m
and its four in-frame children commit 61.9m between them.

**It is not removed here.** This document extracts; the exclusion is an
analytical decision and belongs where it can be re-run without the 32 GB
source. What this section exports is the parent-child map, a few kilobytes, so
that `construction/` can apply the filter — and so that a reader asking "what
if the children are kept?" can answer it by editing one line of R rather than
re-downloading the corpus.

```{sql, connection = con}
DROP TABLE IF EXISTS gbs_related;
```

```{sql, connection = con}
CREATE TABLE gbs_related AS
SELECT DISTINCT ra.iatiidentifier AS child_id, ra.ref AS parent_id
FROM relatedactivity ra
WHERE ra.type = '1'
  AND ra.iatiidentifier IN (SELECT iatiidentifier FROM gbs_frame)
  AND ra.ref            IN (SELECT iatiidentifier FROM gbs_frame);
```

```{sql, connection = con}
SELECT COUNT(*) AS pairs,
       COUNT(DISTINCT child_id)  AS children,
       COUNT(DISTINCT parent_id) AS parents
FROM gbs_related;
```

## 2. Does the frame clear the design thresholds?

The pre-registered threshold for a third level in the model is 20 donors. The
purpose-code frame returned 15.

```{sql, connection = con}
SELECT COUNT(*) AS n_activities, COUNT(DISTINCT reportingorg_ref) AS n_donors
FROM gbs_frame;
```

The World Bank test. The purpose-code frame returned 5 activities for the
largest budget support provider in the world; if undercoverage is the right
diagnosis, this returns materially more.

```{sql, connection = con}
SELECT reportingorg_ref, MIN(reportingorg_narrative) AS donor, COUNT(*) AS n
FROM gbs_frame
WHERE reportingorg_narrative LIKE '%world bank%'
   OR reportingorg_narrative LIKE '%international development association%'
   OR reportingorg_ref LIKE '44000%'
GROUP BY reportingorg_ref
ORDER BY n DESC;
```

Which convention each publisher uses, and how many activities consulting only
the activity-level element would have lost.

```{sql, connection = con}
SELECT declared_activity_level, declared_transaction_level,
       COUNT(*) AS n_activities, COUNT(DISTINCT reportingorg_ref) AS n_donors
FROM gbs_frame
GROUP BY declared_activity_level, declared_transaction_level;
```

```{sql, connection = con}
SELECT reportingorg_ref, MIN(reportingorg_narrative) AS donor, reportingorg_type,
       COUNT(*) AS n_activities, SUM(has_a01) AS n_a01, SUM(has_a02) AS n_a02
FROM gbs_frame
GROUP BY reportingorg_ref, reportingorg_type
ORDER BY n_activities DESC;
```

## 3. Temporal window

2014 onward: IATI reporting is irregular before this date, which is a property
of the source and therefore belongs here.

**No upper bound is imposed.** The analytical window closes in 2023 because
the aid-dependence control is published with a two-year lag — but that is a
fact about the World Bank's release schedule, not about IATI, and it will
change when 2024 figures appear. Encoding it here would mean re-downloading 32
GB to move the window. It is applied in construction/01_build_tables, as a
constant at the top of the file.

```{sql, connection = con}
DROP TABLE IF EXISTS gbs_frame_windowed;
```

```{sql, connection = con}
CREATE TABLE gbs_frame_windowed AS
SELECT f.*,
       MIN(t.transactiondate_isodate) AS first_txn,
       MAX(t.transactiondate_isodate) AS last_txn,
       COUNT(*)                       AS n_transactions
FROM gbs_frame f
JOIN transaction_breakdown t ON t._link_activity = f.act
WHERE t.transactiondate_isodate >= '2014-01-01'
GROUP BY f.act;
```

```{sql, connection = con}
SELECT COUNT(*) AS n_activities, COUNT(DISTINCT reportingorg_ref) AS n_donors
FROM gbs_frame_windowed;
```

## 4. Is the planned-disbursement schedule available?

The outcome originally specified — deviation from the donor's own published
schedule — depends entirely on this field being populated.

```{sql, connection = con}
SELECT f.reportingorg_ref,
       MIN(f.reportingorg_narrative) AS donor,
       COUNT(DISTINCT f.act)         AS n_activities,
       COUNT(DISTINCT p._link_activity) AS n_with_planned,
       ROUND(100.0 * COUNT(DISTINCT p._link_activity)
                   / NULLIF(COUNT(DISTINCT f.act), 0)) AS pct_with_planned
FROM gbs_frame_windowed f
LEFT JOIN planneddisbursement p ON p._link_activity = f.act
GROUP BY f.reportingorg_ref
ORDER BY n_activities DESC;
```

## 5. Document links, paired with their categories

The flat CSV export could not deliver this: repeated child elements are
comma-joined there, and for 628 of 1,118 activities the URL count did not match
the category count, so re-pairing by position would have misfiled document types
across most of the corpus. Here the relation is native.

```{sql, connection = con}
DROP TABLE IF EXISTS gbs_documents;
```

```{sql, connection = con}
CREATE TABLE gbs_documents AS
SELECT f.iatiidentifier, f.reportingorg_ref, f.reportingorg_narrative,
       dl.url, dl.format, dc.code AS category_code
FROM gbs_frame_windowed f
JOIN documentlink dl ON dl._link_activity = f.act
LEFT JOIN documentlink_category dc ON dc._link_documentlink = dl._link;
```

```{sql, connection = con}
SELECT category_code, COUNT(*) AS n_links,
       COUNT(DISTINCT iatiidentifier) AS n_activities
FROM gbs_documents
GROUP BY category_code
ORDER BY n_links DESC;
```

## 6. Export

```{r}
frame <- dbGetQuery(con, "SELECT * FROM gbs_frame_windowed") |> as_tibble()
docs  <- dbGetQuery(con, "SELECT * FROM gbs_documents WHERE url IS NOT NULL") |> as_tibble()

# Harvest runtime is bounded by per-domain crawl delay, so the domain split is
# what tells us whether the download is hours or days.
docs <- docs |>
  mutate(domain = str_extract(url, "(?<=://)[^/]+"))

docs |> count(domain, sort = TRUE) |> print(n = 20)

write_csv(frame, "../data/interim/gbs_frame.csv")
write_csv(docs,  "../data/interim/gbs_documents.csv")
```

## 7. Raw ingredients for `construction/`

Everything below stays in this document, not in `construction/`, for one reason:
this is the only script in the repository that touches the 32 GB source. Every
field `construction/` needs — dates, budget, recipients, per-year transaction
totals, the activity-level controls — is pulled here, once, and written to a CSV
that is committed. `construction/` then runs end to end on a clean clone with no
access to `iati.sqlite`, only to what this section exports.

Recipients, with their reported percentage — the source for the single-country
assignment built downstream.

```{sql, connection = con}
DROP TABLE IF EXISTS gbs_recipients;
```

```{sql, connection = con}
CREATE TABLE gbs_recipients AS
SELECT f.iatiidentifier, rc.code AS country, rc.percentage
FROM gbs_frame_windowed f
JOIN recipientcountry rc ON rc._link_activity = f.act;
```

Activity dates. IATI encodes the four date types two ways across publishers —
numeric codes (`1`-`4`) and string codes (`start-actual`, `end-planned`, ...) —
so both are mapped to one label before pivoting.

```{sql, connection = con}
DROP TABLE IF EXISTS gbs_dates;
```

```{sql, connection = con}
CREATE TABLE gbs_dates AS
SELECT f.iatiidentifier,
       CASE d.type
         WHEN '1' THEN 'planned_start' WHEN 'start-planned' THEN 'planned_start'
         WHEN '2' THEN 'actual_start'  WHEN 'start-actual'  THEN 'actual_start'
         WHEN '3' THEN 'planned_end'   WHEN 'end-planned'   THEN 'planned_end'
         WHEN '4' THEN 'actual_end'    WHEN 'end-actual'    THEN 'actual_end'
       END AS date_type,
       d.isodate
FROM gbs_frame_windowed f
JOIN activitydate d ON d._link_activity = f.act
WHERE d.isodate IS NOT NULL AND d.isodate != '';
```

Forward budget. Restricted to `USD`-denominated lines: only 57 per cent of all
budget rows in the corpus report in USD, and a sum mixing currencies without
conversion would be meaningless. This is a real coverage loss, not a
simplification, and is reported as such rather than patched with an unsourced
FX table.

```{sql, connection = con}
DROP TABLE IF EXISTS gbs_budget;
```

```{sql, connection = con}
CREATE TABLE gbs_budget AS
SELECT f.iatiidentifier,
       SUM(b.value)       AS budget_orig,
       COUNT(*)           AS n_budget_lines,
       b.value_currency   AS budget_ccy
FROM gbs_frame_windowed f
JOIN budget b ON b._link_activity = f.act
WHERE b.value_currency = 'USD'
GROUP BY f.iatiidentifier, b.value_currency;
```

Commitment and disbursement transactions (types 2 and 3), in USD. Two exports:
one **windowed** to the frame's temporal frame (2014 onward, matching the
frame construction in Section 3), one **unwindowed** covering the activity's
whole recorded life. The two diverge exactly where a commitment predates the
window — the reason `construction/` needs both rather than deriving one from
the other.

```{sql, connection = con}
DROP TABLE IF EXISTS gbs_tx;
```

```{sql, connection = con}
CREATE TABLE gbs_tx AS
SELECT f.iatiidentifier,
       t.transactiontype_code AS ttype,
       SUBSTR(t.transactiondate_isodate, 1, 4) AS year,
       t.transactiondate_isodate AS tdate,
       t.value_usd AS usd
FROM gbs_frame_windowed f
JOIN transaction_breakdown t ON t._link_activity = f.act
WHERE t.transactiontype_code IN ('2', '3');
```

```{sql, connection = con}
DROP TABLE IF EXISTS gbs_lifecycle;
```

```{sql, connection = con}
CREATE TABLE gbs_lifecycle AS
SELECT iatiidentifier,
       SUM(CASE WHEN ttype = '2' THEN usd ELSE 0 END) AS commitment_usd_full,
       SUM(CASE WHEN ttype = '3' THEN usd ELSE 0 END) AS disbursement_usd_full,
       COUNT(DISTINCT CASE WHEN ttype = '3' THEN year END) AS n_disb_years_full
FROM gbs_tx
GROUP BY iatiidentifier;
```

Activity-level controls, read directly off `activity`. The `_codename` columns
travel with the codes: IATI ships the official label for each code, so the
labels used downstream are authoritative rather than transcribed by hand. This
also surfaces publisher errors — three `finance_type` values in the frame
resolve to "GNI: Gross National Income", "ODA % GNI" and "Population", which
belong to a different codelist entirely.

```{sql, connection = con}
DROP TABLE IF EXISTS gbs_controls;
```

```{sql, connection = con}
CREATE TABLE gbs_controls AS
SELECT f.iatiidentifier,
       a.defaultfinancetype_code       AS finance_type,
       a.defaultfinancetype_codename   AS finance_type_name,
       a.defaultflowtype_code          AS flow_type,
       a.defaultflowtype_codename      AS flow_type_name,
       a.defaulttiedstatus_code        AS tied_status,
       a.collaborationtype_code        AS collaboration_type,
       a.collaborationtype_codename    AS collaboration_type_name,
       a.hierarchy
FROM gbs_frame_windowed f
JOIN activity a ON a._link = f.act;
```

```{r}
recipients <- dbGetQuery(con, "SELECT * FROM gbs_recipients") |> as_tibble()
dates      <- dbGetQuery(con, "SELECT * FROM gbs_dates") |> as_tibble()
budget     <- dbGetQuery(con, "SELECT * FROM gbs_budget") |> as_tibble()
tx         <- dbGetQuery(con, "SELECT * FROM gbs_tx") |> as_tibble()
lifecycle  <- dbGetQuery(con, "SELECT * FROM gbs_lifecycle") |> as_tibble()
controls   <- dbGetQuery(con, "SELECT * FROM gbs_controls") |> as_tibble()
related    <- dbGetQuery(con, "SELECT * FROM gbs_related") |> as_tibble()

write_csv(recipients, "../data/interim/gbs_recipients.csv")
write_csv(dates,      "../data/interim/gbs_dates.csv")
write_csv(budget,     "../data/interim/gbs_budget.csv")
write_csv(tx,         "../data/interim/gbs_tx.csv")
write_csv(lifecycle,  "../data/interim/gbs_lifecycle.csv")
write_csv(controls,   "../data/interim/gbs_controls.csv")
write_csv(related,    "../data/interim/gbs_related.csv")

dbDisconnect(con)
```

**Everything `construction/` needs is now in `data/interim/` as committed CSVs.**
No script past this point opens `iati.sqlite`.
