---
title: "Document harvest"
subtitle: "Downloading the programme documents behind the IATI document links"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: show
execute:
  warning: false
  message: false
  eval: false
---

Drives the harvester in `_harvest_functions.R`. `eval: false` by design: a full
run takes hours and must not fire on every render. Run the chunks interactively.

Two passes were made, differing only in which document categories they target.
Set `CATEGORIES` and re-run.

The first pass covers the whole frame and pulls categories A01, A04 and A08 —
conditions as demanded, from appraisal, legal agreement and results framework
documents. The second narrows to the analytical sample and pulls A05 and A07 —
conditions as resolved, from implementation review and evaluation documents.

The second pass exists because the first could not answer the question. A01/A04/A08
are drafted before or during an operation and record intent; whether a tranche was
actually released would appear in implementation and completion reporting. That
turned out not to be there either, which is a finding in its own right.

```{r}
library(tidyverse)
library(DBI)
library(RSQLite)

source("_harvest_functions.R")

CATEGORIES <- c("A05", "A07")

# Pass 2 restricts to activities with a computable outcome: documents we cannot
# cross-reference against a disbursement record are not worth the crawl delay.
RESTRICT_TO_SAMPLE <- TRUE
```

## 1. What is available

```{r}
con <- dbConnect(SQLite(), DB_PATH)

dbGetQuery(con, "
  SELECT category, COUNT(DISTINCT url) AS n_urls,
         COUNT(DISTINCT iatiidentifier) AS n_activities
  FROM documents WHERE url IS NOT NULL
  GROUP BY category ORDER BY n_urls DESC")

ids <- if (RESTRICT_TO_SAMPLE) {
  dbGetQuery(con, "
    SELECT iatiidentifier FROM analysis
    WHERE complete_flag = 1 AND y1_gap_full IS NOT NULL")$iatiidentifier
} else {
  NULL
}

dbDisconnect(con)
```

## 2. Build the download list

```{r}
links <- build_document_index(categories = CATEGORIES, ids = ids)

links |> count(domain, sort = TRUE)

# Runtime is bounded by politeness, not bandwidth: the AfDB's robots.txt sets
# Crawl-delay 10 and the harvester is sequential by design.
links |>
  mutate(delay = map_dbl(domain, .delay_for)) |>
  summarise(n_urls = n(), hours = sum(delay) / 3600)
```

## 3. Smoke test

Twenty-five documents exercise every resolver — AfDB unwrapping, the World Bank
API, the EC landing-page hop — and surface a broken selector in two minutes
rather than at hour six.

```{r}
harvest_documents(links, limit = 25)
harvest_report()
```

Check `pct_ok` per domain before continuing. A domain returning `not_pdf` across
the board means its resolver needs work, not that its documents are missing.

## 4. Full run

Resumable. If interrupted, re-run this chunk: anything already resolved is
skipped.

```{r}
harvest_documents(links)
harvest_report()
```

## 5. Coverage by donor

A donor with no retrievable documents cannot enter a text-based measure at all,
and that exclusion has to be reported rather than absorbed.

```{r}
con <- dbConnect(SQLite(), DB_PATH)
frame <- dbGetQuery(con, "SELECT iatiidentifier, donor FROM analysis")
dbDisconnect(con)

manifest_read() |>
  mutate(ok = status %in% c("ok", "cached")) |>
  group_by(iati_identifier) |>
  summarise(got_pdf = any(ok), .groups = "drop") |>
  right_join(frame, by = c("iati_identifier" = "iatiidentifier")) |>
  group_by(donor) |>
  summarise(
    n_activities = n(),
    with_pdf     = sum(got_pdf, na.rm = TRUE),
    pct          = round(100 * mean(got_pdf, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  arrange(desc(n_activities))
```

## 6. Failures

```{r}
manifest_read() |>
  filter(!status %in% c("ok", "cached")) |>
  count(domain, status, sort = TRUE)
```

`error` and timeouts are transient and clear on a re-run of section 4. `not_pdf`,
`http_404` and `unresolvable` are terminal and are never retried — if a whole
domain lands there, fix its resolver in `_harvest_functions.R` and delete those
rows from `data/interim/document_manifest.csv` first.
