---
title: "EDA I — The dataset before any cleaning"
author: "Gaspar Vigneaux"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
execute:
  warning: false
  message: false
---

Two jobs, in this order.

**First**, understand every variable: what it contains, what is missing, and whether
it measures what it is supposed to measure. Nothing is cleaned here — problems are
found and named, and the fixes go to 02_feature_engineering.

**Second**, decide the enforcement measure, applying five criteria fixed
before any of this was inspected:

1. **Theoretical fit** — withholding is a decision at a point in time, not a lifetime residual
2. **Convergent validity** — must move with early closure in the direction enforcement implies
3. **Donor coverage** — must not collapse the sample below what the comparison needs
4. **Discrimination** — must distribute, not concentrate in one mode
5. **Falsifiability** — if nothing passes (2), that is the finding, and modelling stops

```{r}
library(tidyverse)
library(DataExplorer)
library(scales)
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")

analysis   <- read_csv("../data/interim/analysis.csv", show_col_types = FALSE)
panel_year <- read_csv("../data/interim/panel_year.csv", show_col_types = FALSE)
structural <- read_csv("../data/interim/structural.csv", show_col_types = FALSE)
controls   <- read_csv("../data/external/controls_country_year.csv", show_col_types = FALSE)
political  <- read_csv("../data/external/controls_political.csv", show_col_types = FALSE)
```

```{r}
#| echo: false
# Every figure quoted in the prose below is computed here, so a changed frame
# updates the text instead of leaving it asserting stale numbers.
pc  <- function(x) round(100 * mean(x, na.rm = TRUE))
pm  <- function(v) round(100 * mean(is.na(analysis[[v]])))
cnt <- function(v, val) sum(analysis[[v]] == val, na.rm = TRUE)
```

# Part I — What is in the data

## 1. Structure and missingness

```{r}
plot_intro(analysis)
introduce(analysis)
```

```{r}
analysis |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  mutate(pct = round(100 * n_missing / nrow(analysis))) |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing)) |>
  print(n = 30)
```

- **`budget_orig` and its two companions are missing for `r pm("budget_orig")`%** of
  activities. The extraction restricts budget lines to those denominated in USD,
  because summing mixed currencies without conversion is meaningless — the
  missingness is that restriction, not absent reporting.
- **`y2_early_days` is missing for `r pm("y2_early_days")`%**, which matters more
  than it looks: this is the convergent-validity instrument used in Part II. It is
  missing wherever an activity has no recorded actual end date, i.e. mostly
  activities still running.
- **`finance_type` and `flow_type` are missing for `r pm("finance_type")`% and
  `r pm("flow_type")`%**, and both were intended as controls. At that rate they
  cannot enter a model without either dropping a third of the sample or an explicit
  missingness treatment.
- **The windowed variables are far sparser than the full-life ones**: `y1_gap` is
  missing for `r pm("y1_gap")`% against `r pm("y1_gap_full")`% for `y1_gap_full`,
  and `commitment_usd` for `r pm("commitment_usd")`% against
  `r pm("commitment_usd_full")`%. This follows directly from closing the window at
  2023 — an activity whose commitment falls outside 2014-2023 has no windowed
  commitment, so the windowed ratio is undefined. The full-life versions are the
  ones to rely on.

## 2. Categorical variables

```{r}
cat_vars <- c("x1_donor_type", "modality", "status_name", "finance_type",
              "flow_type", "tied_status", "collaboration_type", "hierarchy")

analysis |>
  select(all_of(cat_vars)) |>
  mutate(across(everything(), as.character)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  count(variable, value) |>
  mutate(pct = round(100 * n / sum(n)), .by = variable) |>
  arrange(variable, desc(n)) |>
  print(n = 40)
```

```{r}
#| label: fig-categoricals
#| fig-cap: "Distribution of the activity-level categorical variables"
analysis |>
  select(x1_donor_type, modality, status_name, hierarchy) |>
  mutate(across(everything(), as.character)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  count(variable, value) |>
  mutate(value = fct_reorder(value, n)) |>
  ggplot(aes(n, value)) +
  geom_col(fill = "grey30") +
  facet_wrap(~variable, scales = "free") +
  labs(x = "Activities", y = NULL) +
  theme_minimal(base_size = 10)
```

- **The sample is `r pc(analysis$x1_donor_type == "Multilateral")`% multilateral**
  (`r cnt("x1_donor_type","Multilateral")` against
  `r cnt("x1_donor_type","Bilateral")` bilateral). This is the central imbalance of
  the design: the research question compares institutional types, and one type
  carries four fifths of the observations.
- **Sector budget support outnumbers general**
  (`r cnt("modality","A02_sector")` to `r cnt("modality","A01_general")`), which
  follows from the World Bank reporting its Development Policy Financing
  exclusively as A02. Seventeen activities declare both and are kept as their own
  category rather than forced into one.
- **`tied_status` is effectively constant** — `r cnt("tied_status",5)` of
  `r sum(!is.na(analysis$tied_status))` non-missing activities are code 5, untied.
  A variable with no variation cannot be a control, and this one should be dropped
  rather than carried forward.
- **`hierarchy` includes an undocumented level 0** on `r cnt("hierarchy",0)`
  activities. The IATI standard defines 1 (activity) and 2 (sub-activity); 0 is not
  in the codelist and needs checking before use. Level 2 now holds only
  `r cnt("hierarchy",2)` activities — most sub-activities were removed as
  parent-child duplicates in construction/01_build_tables, which is that
  exclusion working as intended.
- `finance_type` splits mainly between 110 (standard grant,
  `r cnt("finance_type",110)`) and 421 (standard loan, `r cnt("finance_type",421)`)
  — a genuine and usable distinction, if the missingness is handled.

## 3. Numeric variables

```{r}
num_vars <- c("commitment_usd_full", "disbursement_usd_full", "n_disb_years_full",
              "planned_duration_days", "y2_early_days", "budget_orig")

analysis |> select(all_of(num_vars)) |> summary()
```

**Impossible or suspicious values.** Each is counted rather than described, because
each one implies a different fix.

```{r}
analysis |>
  summarise(
    commit_zero   = sum(commitment_usd_full == 0, na.rm = TRUE),
    commit_neg    = sum(commitment_usd_full < 0, na.rm = TRUE),
    disb_neg      = sum(disbursement_usd_full < 0, na.rm = TRUE),
    budget_neg    = sum(budget_orig < 0, na.rm = TRUE),
    duration_neg  = sum(planned_duration_days < 0, na.rm = TRUE),
    duration_over_20y = sum(planned_duration_days > 7300, na.rm = TRUE),
    early_neg     = sum(y2_early_days < 0, na.rm = TRUE)
  ) |>
  pivot_longer(everything(), names_to = "check", values_to = "n")
```

- **`r cnt("commitment_usd_full", 0)` activities report zero full-life commitment**
  and `r sum(analysis$commitment_usd_full < 0, na.rm = TRUE)` report negative. A zero
  commitment makes the gap undefined — a structural consequence of the ratio, not
  missing data.
- **`planned_duration_days` ranges from
  `r min(analysis$planned_duration_days, na.rm = TRUE)` to
  `r format(max(analysis$planned_duration_days, na.rm = TRUE), big.mark = ",")`
  days.** A negative planned duration means the planned end precedes the planned
  start, which cannot be true; the maximum is over a century. Both are reporting
  errors and need a plausibility bound.
- **`r sum(analysis$y2_early_days < 0, na.rm = TRUE)` activities have negative
  `y2_early_days`**, meaning they closed *after* their announced date. That is not an
  error but the other tail of the same variable, and it is informative. The extreme
  is: `r min(analysis$y2_early_days, na.rm = TRUE)` days, thirteen years late.
- **`budget_orig` is negative for `r sum(analysis$budget_orig < 0, na.rm = TRUE)` of
  the `r sum(!is.na(analysis$budget_orig))` activities that have it**, which is
  investigated next because it disqualifies the variable as it stands.

### 3.1 The budget variable is not measuring what it should

```{r}
analysis |>
  filter(!is.na(budget_orig)) |>
  mutate(sign = if_else(budget_orig < 0, "negative", "positive")) |>
  count(donor, sign) |>
  pivot_wider(names_from = sign, values_from = n, values_fill = 0) |>
  arrange(desc(negative)) |>
  print(n = 10)
```

- **Every negative budget belongs to the World Bank.** This is not random noise: it
  is a publisher-specific sign convention, and the World Bank is the largest single
  contributor to the analytical sample.
- The extraction sums all USD budget lines for an activity without separating
  `type` (original vs revised) or `status` (indicative vs committed). Where a
  publisher records revisions as negative adjustments, that sum is meaningless.
- **`budget_orig` cannot be used as constructed.** Either the extraction is
  rewritten to take original-committed lines only, or the variable is dropped. Given
  it is already 75% missing, dropping is the likely answer — recorded for
  02_feature_engineering.

```{r}
#| label: fig-numeric-dist
#| fig-cap: "Distribution of the monetary and duration variables (log scale where skewed)"
analysis |>
  select(commitment_usd_full, disbursement_usd_full, planned_duration_days) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  filter(value > 0) |>
  ggplot(aes(value)) +
  geom_histogram(bins = 40, fill = "grey30") +
  scale_x_log10(labels = label_number(scale_cut = cut_short_scale())) +
  facet_wrap(~variable, scales = "free") +
  labs(x = "value (log scale)", y = "Activities") +
  theme_minimal(base_size = 10)
```

- All three are heavily right-skewed and approximately log-normal once zeros and
  negatives are removed. Any model using them should use the log, which is what
  the Statistics course prescribes for this shape.

## 4. Structural variables

Built in construction/04_structural: S1 the donor's share of the recipient's
budget support, S2 the fragmentation of the pool, S3 its size.

```{r}
structural |> select(s1_share, s2_hhi, s2_hhi_net, s3_n_donors) |> summary()
```

```{r}
structural |>
  summarise(
    donor_country_years = n(),
    single_donor        = sum(s3_n_donors == 1),
    pct_single          = round(100 * mean(s3_n_donors == 1)),
    share_exactly_1     = sum(s1_share > 0.999),
    hhi_net_missing     = sum(is.na(s2_hhi_net))
  )
```

```{r}
#| label: fig-structural
#| fig-cap: "Donor share and pool size across donor-country-years"
structural |>
  select(s1_share, s3_n_donors) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 30, fill = "grey30") +
  facet_wrap(~variable, scales = "free") +
  labs(x = NULL, y = "Donor-country-years") +
  theme_minimal(base_size = 10)
```

- **A third of donor-country-years have a single donor**, and in those S1 is exactly
  1 by construction. That value carries no information about leverage — it is what
  a one-donor cell must produce — so those cells contribute nothing to identifying
  H2 and need to be either excluded or flagged.
- **`s2_hhi_net` is missing for exactly those 687 single-donor cells**, which is
  correct behaviour: a pool net of the only donor in it is empty, not zero. But it
  means H3 is estimable on two thirds of the cells, not all.
- **The donor pool never exceeds 7** and its median is 2. This is a much narrower
  range than the fragmentation literature assumes — Molenaers et al. work with
  recipient countries hosting a dozen or more budget support donors. The
  restricted range limits how much of the fragmentation hypothesis this data can
  test, and that limit belongs in the write-up.

## 5. External controls

```{r}
controls |>
  select(voice_account, corruption_ctrl, pol_stability, aid_dependence, gdp_pc, population) |>
  summary()
```

```{r}
bind_rows(
  controls |> summarise(source = "World Bank", rows = n(), countries = n_distinct(iso2),
                        year_min = min(year), year_max = max(year)),
  political |> summarise(source = "V-Dem", rows = n(), countries = n_distinct(iso2),
                         year_min = min(year), year_max = max(year))
)
```

- **`aid_dependence` is missing for 49% of country-years** — by far the worst
  coverage of any control, and the one Molenaers et al. find most predictive. Net
  ODA as a share of GNI is only published for aid-receiving countries, so the
  missingness is structural: high-income countries in the table have no value
  because the concept does not apply to them. Restricted to actual recipients,
  coverage should be far better, and that is the check to run before deciding.
- The three governance indicators are missing for roughly 13% each, concentrated in
  the most recent years — the WGI are published with a lag and end in 2024.
- **Both sources stop in 2025**, while the frame runs to June 2026. Activity-years
  in 2026 will have no control values, and carrying the last observation forward
  would be inventing a political history that has not happened yet.

```{r}
political |>
  summarise(coup_years = sum(coup), election_years = sum(election_year),
            pct_coup = round(100 * mean(coup)), pct_election = round(100 * mean(election_year)))
```

- 17 coup-years and 602 election-years across the window. Coups are rare enough that
  they will only ever be a sparse indicator; elections, at roughly 29% of
  country-years, have enough variation to function as a control.

## 6. Joining the sources

The three tables key differently — activity, donor-country-year, country-year — and
the join is where coverage is actually lost.

```{r}
analysis |>
  summarise(
    activities            = n(),
    recipient_missing     = sum(is.na(recipient_main)),
    country_in_wb_controls = sum(recipient_main %in% controls$iso2, na.rm = TRUE)
  )

analysis |>
  filter(!is.na(recipient_main), !recipient_main %in% controls$iso2) |>
  count(recipient_main, sort = TRUE)
```

- **11 activities have no recipient country at all** and cannot receive any
  country-level control.
- **85 activities name a country absent from the World Bank tables** — Niue, the
  Cook Islands, Tokelau, Saint Pierre and Miquelon, Saint Barthélemy, Wallis and
  Futuna. These are small Pacific and French overseas territories that the World
  Bank does not report on. They are a genuine coverage loss, concentrated in one
  donor's portfolio, and should be reported rather than dropped silently.

## 7. What Part I found, and what it hands to feature engineering

| Variable | Problem | Likely fix |
|---|---|---|
| `budget_orig` | 75% missing; all 655 negatives are one publisher's sign convention | Drop, or re-extract original-committed lines only |
| `tied_status` | 98% one value | Drop — no variation to exploit |
| `hierarchy` | Undocumented level 0 on 337 activities | Check against the codelist before use |
| `planned_duration_days` | Negative values; maximum of 105 years | Plausibility bounds |
| `y2_early_days` | Extremes to ±13 years | Winsorise or bound; keep the sign, it is informative |
| `commitment_usd_full` | 145 zeros, 2 negatives | Exclude from ratio-based measures |
| Monetary variables | Strong right skew | Log transform |
| `aid_dependence` | 49% missing overall | Re-check restricted to actual recipients |
| `finance_type`, `flow_type` | 28-29% missing | Explicit missing category, or drop |
| S1 in single-donor cells | Exactly 1 by construction, uninformative | Flag or exclude the 33% single-donor cells |
| Controls after 2025 | Source ends before the frame does | Restrict the window; do not carry forward |
| 96 activities | No recipient, or country absent from controls | Report as coverage loss |

# Part II — Choosing the enforcement measure

Everything above is descriptive. This part applies the five criteria and decides.

## 8. The filter cascade

The analytical sample is not the frame. Each restriction removes donors, and which
donors it removes is not random — this is criterion 3, and it has to be counted.

```{r}
cascade <- tibble(
  step = c("Frame", "+ status Finalisation or Closed", "+ computable full-life gap"),
  activities = c(
    nrow(analysis),
    sum(analysis$complete_flag == 1),
    sum(analysis$complete_flag == 1 & !is.na(analysis$y1_gap_full))
  ),
  donors = c(
    n_distinct(analysis$donor_ref),
    n_distinct(analysis$donor_ref[analysis$complete_flag == 1]),
    n_distinct(analysis$donor_ref[analysis$complete_flag == 1 & !is.na(analysis$y1_gap_full)])
  )
)

cascade |> mutate(pct_activities = round(100 * activities / activities[1]))
```

**Why completion is required at all.** An activity still under implementation shows
a large gap for no reason other than that its disbursements have not happened yet:
53% of in-implementation activities exhibit a gap above 0.9. Without the
restriction, more than half the positive cases would be false ones. The restriction
is correct for this measure — and it costs 32 of 51 donors.

```{r}
lost_donors <- analysis |>
  mutate(in_sample = complete_flag == 1 & !is.na(y1_gap_full)) |>
  summarise(n_activities = n(), n_in_sample = sum(in_sample),
            .by = c(donor, x1_donor_type)) |>
  filter(n_in_sample == 0) |>
  arrange(desc(n_activities))

lost_donors |> print(n = 40)
lost_donors |> count(x1_donor_type, wt = n_activities, name = "activities_lost")
```

```{r}
#| label: fig-cascade
#| fig-cap: "Activities and donors surviving each restriction"
cascade |>
  mutate(step = fct_inorder(step)) |>
  pivot_longer(c(activities, donors)) |>
  ggplot(aes(step, value)) +
  geom_col(fill = "grey30") +
  geom_text(aes(label = value), vjust = -0.4, size = 3.5) +
  facet_wrap(~name, scales = "free_y") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

## 9. Candidate A — the full-life gap

The measure specified early in the project and later found wanting, diagnosed
here in full so the rejection rests on visible evidence.

```{r}
gap_data <- analysis |> filter(complete_flag == 1, !is.na(y1_gap_full))

gap_data |>
  summarise(
    n = n(),
    mean = round(mean(y1_gap_full), 3),
    median = round(median(y1_gap_full), 3),
    sd = round(sd(y1_gap_full), 3),
    pct_zero = round(100 * mean(abs(y1_gap_full) < 0.01)),
    pct_neg = round(100 * mean(y1_gap_full < -0.01)),
    pct_above_90 = round(100 * mean(y1_gap_full > 0.9))
  )
```

```{r}
#| label: fig-gap-dist
#| fig-cap: "Distribution of the full-life commitment-disbursement gap"
gap_data |>
  filter(between(y1_gap_full, -0.5, 1.5)) |>
  ggplot(aes(y1_gap_full)) +
  geom_histogram(bins = 60, fill = "grey30") +
  geom_vline(xintercept = 0.25, colour = "firebrick", linetype = "dashed") +
  scale_x_continuous(labels = percent) +
  labs(x = "Gap (1 - disbursed/committed)", y = "Activities",
       caption = "Dashed line: the provisional 0.25 withholding threshold") +
  theme_minimal(base_size = 11)
```

**Criterion 4 — discrimination.** The band between the mass at zero and the mass at
near-total non-disbursement is where a graded measure would carry its information.

```{r}
gap_data |>
  mutate(band = cut(y1_gap_full, c(-Inf, -0.01, 0.01, 0.1, 0.25, 0.5, 0.9, Inf),
                    labels = c("negative", "~zero", "0-10%", "10-25%",
                               "25-50%", "50-90%", ">90%"))) |>
  count(band) |>
  mutate(pct = round(100 * n / sum(n)))
```

- 62% sit within rounding of zero and 14% are negative — the latter are activities
  that disbursed more than they committed, from currency revaluation or
  supplementary financing.
- **Only 4% of activities fall between 10% and 90%.** The middle of the range is
  empty. This is the signature of two distinct populations, not one graded
  construct, and it means a threshold anywhere in that band is cutting through a
  void rather than separating cases.

## 10. Candidate B — withholding as a year-level event

Criterion 1 asks for a measure representing a decision taken at a point in time.
This candidate asks a different question of the same transactions: **in a year when
this activity was live and money remained owed, did any disbursement arrive?**

```{r}
activity_span <- panel_year |>
  summarise(first_year = min(year), last_year = max(year), .by = iatiidentifier)

# An interruption only means something between the first and last observed year:
# before the first there is nothing to withhold, after the last the operation has
# stopped and silence is closure, not enforcement.
year_events <- activity_span |>
  reframe(year = first_year:last_year, .by = iatiidentifier) |>
  left_join(panel_year, by = c("iatiidentifier", "year")) |>
  mutate(
    disbursement_usd = replace_na(disbursement_usd, 0),
    commitment_usd   = replace_na(commitment_usd, 0)
  ) |>
  arrange(iatiidentifier, year) |>
  mutate(
    cum_commit   = cumsum(commitment_usd),
    cum_disb     = cumsum(disbursement_usd),
    # Undisbursed balance carried into this year: only if money was still owed
    # can a zero-disbursement year be read as withholding.
    balance_owed = cum_commit - cum_disb,
    interrupted  = as.integer(disbursement_usd <= 0 & balance_owed > 0),
    .by = iatiidentifier
  )

year_events |>
  summarise(activity_years = n(), activities = n_distinct(iatiidentifier),
            pct_interrupted = round(100 * mean(interrupted)))
```

```{r}
activity_interruptions <- year_events |>
  summarise(n_years = n(), n_interrupted = sum(interrupted),
            any_interrupt = as.integer(sum(interrupted) > 0),
            .by = iatiidentifier)

activity_interruptions |> count(n_interrupted) |> arrange(n_interrupted)
```

- **`r round(100*mean(year_events$interrupted))`% of activity-years are flagged.**
  That is a high base rate for what is meant to be a sanctioning event, and it is
  the main open question about this measure:
  budget support disburses in tranches, sometimes biennial, so a zero-disbursement
  year may reflect the tranche schedule rather than a decision to withhold. The
  measure passes the tests below, but almost certainly captures a mixture of
  enforcement and ordinary disbursement lumpiness. Refining it — requiring more
  than one consecutive year, or weighting by the balance owed — is the first task
  of 02_feature_engineering.

**Criterion 3 — does this candidate recover donors?** The full-life gap requires a
completed activity; an interruption does not.

```{r}
bind_rows(
  analysis |>
    filter(complete_flag == 1, !is.na(y1_gap_full)) |>
    summarise(measure = "A: full-life gap", activities = n(), donors = n_distinct(donor_ref)),
  analysis |>
    inner_join(activity_interruptions, by = "iatiidentifier") |>
    summarise(measure = "B: year-level interruption", activities = n(), donors = n_distinct(donor_ref))
)
```

## 11. Criterion 2 — convergent validity

The eliminatory test. Early closure derives from dates rather than money, so it is
an independent channel on the same construct. If a measure captures
enforcement, activities it flags should close early **more** often, not less.

```{r}
convergent <- analysis |>
  filter(complete_flag == 1, !is.na(y2_early_days)) |>
  left_join(activity_interruptions, by = "iatiidentifier") |>
  mutate(closed_early = as.integer(y2_early_days > 30))

# Candidate A at several thresholds: an earlier pass found the relationship ran
# backwards and strengthened as the threshold tightened. Re-checked on the
# corrected frame.
map_dfr(c(0.10, 0.25, 0.50), \(thr) {
  convergent |>
    filter(!is.na(y1_gap_full)) |>
    mutate(flagged = y1_gap_full > thr) |>
    summarise(threshold = thr, n_flagged = sum(flagged),
              early_if_flagged = round(100 * mean(closed_early[flagged])),
              early_if_not = round(100 * mean(closed_early[!flagged])))
})
```

```{r}
convergent |>
  filter(!is.na(any_interrupt)) |>
  summarise(n_flagged = sum(any_interrupt == 1),
            early_if_flagged = round(100 * mean(closed_early[any_interrupt == 1])),
            early_if_not = round(100 * mean(closed_early[any_interrupt == 0])))
```

```{r}
#| label: fig-convergent
#| fig-cap: "Early-closure rate by whether each candidate flags the activity"
bind_rows(
  convergent |> filter(!is.na(y1_gap_full)) |>
    transmute(measure = "A: gap > 25%", flagged = y1_gap_full > 0.25, closed_early),
  convergent |> filter(!is.na(any_interrupt)) |>
    transmute(measure = "B: any interrupted year", flagged = any_interrupt == 1, closed_early)
) |>
  summarise(pct_early = 100 * mean(closed_early), n = n(), .by = c(measure, flagged)) |>
  ggplot(aes(flagged, pct_early, fill = flagged)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(pct_early), "%")), vjust = -0.4, size = 3.5) +
  facet_wrap(~measure) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = c("grey60", "firebrick")) +
  labs(x = "Measure flags the activity", y = "Closed early (%)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

## 12. Decision

Every figure in this table is computed from the objects above rather than typed
in, so re-running the pipeline on a changed frame cannot leave the prose
asserting numbers the code no longer produces.

```{r}
#| echo: false
cov <- bind_rows(
  analysis |> filter(complete_flag == 1, !is.na(y1_gap_full)) |>
    summarise(m = "A", act = n(), don = n_distinct(donor_ref)),
  analysis |> inner_join(activity_interruptions, by = "iatiidentifier") |>
    summarise(m = "B", act = n(), don = n_distinct(donor_ref))
)
A_act <- cov$act[cov$m == "A"]; A_don <- cov$don[cov$m == "A"]
B_act <- cov$act[cov$m == "B"]; B_don <- cov$don[cov$m == "B"]

cvA <- convergent |> filter(!is.na(y1_gap_full)) |>
  mutate(f = y1_gap_full > 0.25) |>
  summarise(hit = round(100*mean(closed_early[f])), miss = round(100*mean(closed_early[!f])))
cvB <- convergent |> filter(!is.na(any_interrupt)) |>
  summarise(hit = round(100*mean(closed_early[any_interrupt == 1])),
            miss = round(100*mean(closed_early[any_interrupt == 0])))

gp <- gap_data |> summarise(z = round(100*mean(abs(y1_gap_full) < 0.01)),
                            n = round(100*mean(y1_gap_full < -0.01)),
                            mid = round(100*mean(between(y1_gap_full, 0.1, 0.9))))
max_int <- max(activity_interruptions$n_interrupted)
```

| Criterion | A: full-life gap | B: year-level interruption |
|---|---|---|
| 1. Theoretical fit | **Fails.** A lifetime residual cannot separate a withholding decision from restructuring or administrative closure | **Passes.** A live year with money owed and nothing disbursed is a decision at a point in time |
| 2. Convergent validity | **Fails.** Flagged activities close early *less* often (`r cvA$hit`% vs `r cvA$miss`%), and the gap widens as the threshold tightens | **Passes.** Flagged activities close early *more* often (`r cvB$hit`% vs `r cvB$miss`%) |
| 3. Donor coverage | **Fails.** `r format(A_act, big.mark=",")` activities, `r A_don` donors — below the pre-registered threshold of 20 | **Passes.** `r format(B_act, big.mark=",")` activities, `r B_don` donors |
| 4. Discrimination | **Fails.** `r gp$z`% within rounding of zero, `r gp$n`% negative, only `r gp$mid`% between 10% and 90% | **Passes.** Interruption counts spread from 0 to `r max_int` years |

**Decision: Candidate B**, subject to the refinement flagged in §10. The measure
carried forward is the year-level interruption — an activity-year in which the
operation was live, an undisbursed balance remained, and no disbursement arrived.

Three consequences, each substantive rather than technical.

**The earlier failure replicates on the corrected frame.** The gap's convergent-validity
failure was first found on the earlier, undocumented pipeline. It reproduces here at
every threshold on the rebuilt frame, which rules out frame construction as its cause.

**The donor ceiling was a property of the measure, not of the data.** An
earlier pass concluded the 14-donor ceiling was structural. That needs
amending: it was structural *to the full-life gap*, which requires a completed
activity. An interruption does not, and the sample reaches **`r B_don`
donors** — above the pre-registered threshold of 20 for the first time in the
project.

**The unit of analysis changes to the activity-year.** This aligns the outcome with
the theory, which specifies a decision taken when a disbursement is due, and with the
structural variables, already computed per donor-country-year.

The fallback does not apply: a candidate satisfied convergent validity, so
modelling proceeds.
