---
title: "Feature engineering"
subtitle: "Cleaning the variables, and refining the enforcement measure"
author: "Gaspar Vigneaux"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
execute:
  warning: false
  message: false
---

Two jobs, both driven by what 01_eda_raw found.

**Part I** applies the fixes listed in that document's closing table — every
variable that was found not to measure what it should.

**Part II** refines the enforcement measure. 01_eda_raw chose the
year-level interruption on the selection criteria, but flagged one problem: at a
base rate near 40 per cent of activity-years, it almost certainly mixes genuine
withholding with the ordinary lumpiness of tranche disbursement. Candidate
refinements are tested against the same criteria that selected the measure —
this is not a fresh search.

```{r}
library(tidyverse)
library(scales)
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")

analysis   <- read_csv("../data/interim/analysis.csv", show_col_types = FALSE)
panel_year <- read_csv("../data/interim/panel_year.csv", show_col_types = FALSE)
structural <- read_csv("../data/interim/structural.csv", show_col_types = FALSE)
controls   <- read_csv("../data/external/controls_country_year.csv", show_col_types = FALSE)
political  <- read_csv("../data/external/controls_political.csv", show_col_types = FALSE)
```

# Part I — Cleaning

## 1. Variables dropped

Two are removed outright, each for a reason established in the EDA rather than
for convenience.

```{r}
# budget_orig: 71% missing, and every negative value belongs to the World Bank —
# a publisher-specific sign convention in revised budget lines, not noise. The
# variable cannot be repaired without re-extracting original-committed lines
# only, and at 71% missing it would not earn its place even then.
# tied_status: 89% a single value (untied). A control with no variation
# identifies nothing; that budget support is untied is expected, not a finding.
dropped <- c("budget_orig", "n_budget_lines", "budget_ccy", "tied_status")

clean <- analysis |> select(-all_of(dropped))

tibble(dropped = dropped)
```

## 2. Implausible values bounded

```{r}
# A planned end before the planned start is impossible; a planned duration
# beyond 20 years is not a budget support operation. Both become NA rather than
# being winsorised: they are reporting errors, not extreme true values.
MAX_DURATION_DAYS <- 7300

# Early closure is genuine information in both tails — closing late is as
# meaningful as closing early — so this is winsorised rather than nulled, to
# keep the sign while removing leverage from a handful of extremes.
EARLY_BOUND_DAYS <- 730

clean <- clean |>
  mutate(
    planned_duration_days = if_else(
      planned_duration_days < 0 | planned_duration_days > MAX_DURATION_DAYS,
      NA_integer_, planned_duration_days
    ),
    y2_early_days_raw = y2_early_days,
    y2_early_days = pmin(pmax(y2_early_days, -EARLY_BOUND_DAYS), EARLY_BOUND_DAYS)
  )

clean |>
  summarise(
    duration_nulled = sum(is.na(planned_duration_days)) - sum(is.na(analysis$planned_duration_days)),
    early_winsorised = sum(y2_early_days != y2_early_days_raw, na.rm = TRUE)
  )
```

## 3. Recoding

```{r}
# hierarchy 0 is not in the IATI codelist. It is concentrated in the IADB and
# a few bilaterals, and marks a standalone operation rather than a component of
# one, which is what level 1 means. Recoded to 1.
#
# finance_type and flow_type are missing for about a third of activities.
# Whether a publisher declares them is plausibly a fact about its reporting
# practice, not about the activity, so the missingness is kept as a level
# rather than imputed away.
clean <- clean |>
  mutate(
    hierarchy = if_else(hierarchy == 0, 1, hierarchy),
    # The codes stay numeric; the labels get the explicit level. IATI's own
    # `_codename` values are used rather than transcribed ones, so the labels
    # are authoritative — including where they reveal a publisher error:
    # finance_type 1, 2 and 4 resolve to "GNI", "ODA % GNI" and "Population",
    # which belong to a different codelist and are flagged rather than recoded.
    across(any_of(c("finance_type_name", "flow_type_name",
                    "collaboration_type_name")),
           ~ fct_na_value_to_level(factor(.x), level = "not declared"))
  )

clean |> count(hierarchy)
if ("finance_type_name" %in% names(clean)) {
  clean |> count(finance_type, finance_type_name, sort = TRUE)
  clean |> count(flow_type, flow_type_name, sort = TRUE)
} else {
  clean |> count(finance_type, sort = TRUE)
}
```

## 4. Log transforms

```{r}
# The monetary variables are approximately log-normal once zeros and negatives
# are removed (EDA §3). log1p on the positive part keeps zeros at zero rather
# than dropping them.
clean <- clean |>
  mutate(
    # if_else evaluates both branches, so log() must not see the non-positive
    # values even though they are discarded — hence the guarded input.
    log_commitment   = if_else(commitment_usd_full > 0,
                               log(pmax(commitment_usd_full, 1)), NA_real_),
    log_disbursement = if_else(disbursement_usd_full > 0,
                               log(pmax(disbursement_usd_full, 1)), NA_real_)
  )
```

## 5. Coverage exclusions

```{r}
# 9 activities carry no recipient country and can receive no country-level
# control. A further set names countries the World Bank does not report on —
# small Pacific and French overseas territories. Both are coverage losses to
# report, not to patch.
before <- nrow(clean)

clean <- clean |>
  filter(!is.na(recipient_main), recipient_main %in% controls$iso2)

tibble(
  activities_before = before,
  dropped_no_country_coverage = before - nrow(clean),
  activities_after = nrow(clean)
)
```

# Part II — Refining the enforcement measure

## 6. The baseline measure

As defined in 01_eda_raw: an activity-year in which the operation was
live, an undisbursed balance remained, and no disbursement arrived.

```{r}
activity_span <- panel_year |>
  filter(iatiidentifier %in% clean$iatiidentifier) |>
  summarise(first_year = min(year), last_year = max(year), .by = iatiidentifier)

year_events <- activity_span |>
  reframe(year = first_year:last_year, .by = iatiidentifier) |>
  left_join(panel_year, by = c("iatiidentifier", "year")) |>
  mutate(
    disbursement_usd = replace_na(disbursement_usd, 0),
    commitment_usd   = replace_na(commitment_usd, 0)
  ) |>
  arrange(iatiidentifier, year) |>
  mutate(
    cum_commit   = cumsum(commitment_usd),
    cum_disb     = cumsum(disbursement_usd),
    balance_owed = cum_commit - cum_disb,
    # Share of the activity's committed money still outstanding. A balance of
    # 2% left over is not the same event as a balance of 80%.
    balance_share = if_else(cum_commit > 0, balance_owed / cum_commit, NA_real_),
    interrupted   = as.integer(disbursement_usd <= 0 & balance_owed > 0),
    .by = iatiidentifier
  )

year_events |>
  summarise(activity_years = n(), activities = n_distinct(iatiidentifier),
            base_rate = round(100 * mean(interrupted)))
```

## 7. Candidate refinements

Three variants, each targeting the same suspicion: that a single
zero-disbursement year reflects a biennial tranche schedule rather than a
decision to withhold.

```{r}
year_events <- year_events |>
  arrange(iatiidentifier, year) |>
  mutate(
    prev_interrupted = lag(interrupted, default = 0L),
    next_interrupted = lead(interrupted, default = 0L),
    # V1 — the interruption persists: a year that is interrupted and adjacent
    # to another interrupted year. A one-off gap is treated as scheduling.
    int_consecutive = as.integer(interrupted == 1 &
                                  (prev_interrupted == 1 | next_interrupted == 1)),
    # V2 — the money at stake is substantial: at least a quarter of committed
    # funds still outstanding. Withholding 2% of a nearly-complete operation is
    # not the sanction the theory describes.
    # balance_share is NA where no commitment has yet been recorded in the
    # window; %in% TRUE collapses that to FALSE rather than propagating NA
    # through the whole variant.
    int_substantial = as.integer((interrupted == 1 & balance_share >= 0.25) %in% TRUE),
    # V3 — both conditions.
    int_strict = as.integer((int_consecutive == 1 & int_substantial == 1) %in% TRUE),
    .by = iatiidentifier
  )

variants <- c(baseline = "interrupted", consecutive = "int_consecutive",
              substantial = "int_substantial", strict = "int_strict")

map_dfr(names(variants), \(v) {
  col <- variants[[v]]
  year_events |>
    summarise(variant = v,
              base_rate = round(100 * mean(.data[[col]])),
              activity_years_flagged = sum(.data[[col]]))
})
```

## 8. The same test that selected the measure

Convergent validity against early closure — the second selection criterion,
applied identically. A refinement earns its place only if it holds or improves the
direction the baseline established.

```{r}
early <- clean |>
  filter(!is.na(y2_early_days)) |>
  transmute(iatiidentifier, closed_early = as.integer(y2_early_days > 30))

convergent_test <- map_dfr(names(variants), \(v) {
  col <- variants[[v]]
  year_events |>
    summarise(flagged = as.integer(sum(.data[[col]]) > 0), .by = iatiidentifier) |>
    inner_join(early, by = "iatiidentifier") |>
    summarise(
      variant          = v,
      n_flagged        = sum(flagged),
      early_if_flagged = round(100 * mean(closed_early[flagged == 1])),
      early_if_not     = round(100 * mean(closed_early[flagged == 0])),
      margin           = early_if_flagged - early_if_not
    )
})

convergent_test
```

```{r}
#| label: fig-variants
#| fig-cap: "Convergent-validity margin and base rate, by candidate refinement"
convergent_test |>
  left_join(
    map_dfr(names(variants), \(v) tibble(variant = v,
      base_rate = 100 * mean(year_events[[variants[[v]]]]))),
    by = "variant"
  ) |>
  ggplot(aes(base_rate, margin, label = variant)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(size = 3, colour = "grey20") +
  geom_text(vjust = -0.9, size = 3.2) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(x = "Base rate (% of activity-years flagged)",
       y = "Convergent-validity margin (pp)",
       caption = "Upper-left is better: a plausible base rate with a wide margin in the predicted direction") +
  theme_minimal(base_size = 11)
```

## 9. Selection

```{r}
#| echo: false
best <- convergent_test |> slice_max(margin, n = 1)
```

The variant carried forward is **`r best$variant`** — an activity-year with no
disbursement in which at least a quarter of committed funds remained
outstanding — with a convergent-validity margin of `r best$margin` percentage
points in the direction enforcement implies.

**What the comparison shows.** The discriminating filter is the *amount at
stake*, not persistence over time. Requiring the interruption to last two
consecutive years produces the weakest margin of the four (+3) while discarding
most of the flagged cases; requiring a substantial outstanding balance produces
the strongest (+10). The initial suspicion — that single zero-disbursement years
were artefacts of a biennial tranche schedule — is not what the data support.
What separates a meaningful interruption from a trivial one is whether real
money was still owed, which is why an operation that disbursed 98 per cent and
left a residual is correctly excluded while one that left 40 per cent unpaid is
not.

**Two alternative operationalisations were tested and rejected**, on the same
criterion rather than on judgement. Deviation from an activity's own
disbursement rhythm returns a margin of **−8**, the wrong direction entirely;
an unusually long interval since the last disbursement flags nothing at all.
Both fail for the same reason: only 273 of 2,213 activities span four or more
years, so there is no established rhythm to deviate from. The panel is too
short per activity to support a within-activity relative measure, and that is a
property of the data rather than of the idea.

**Naming, and what is being claimed.** The variable is called `interrupted`
rather than `withheld`. The theoretical construct is withholding — a donor's
decision not to disburse when a benchmark is missed — but the measure observes
an interruption, and not every interruption is a sanction. The convergent
validity test establishes that the set correlates with an independent signal of
enforcement; it does not establish that each flagged year is a sanction.
Claiming the latter would overstate what the data can carry. The measure is an
observable proxy for the construct and is described as one throughout.

```{r}
OUTCOME_VAR <- best$variant

year_events <- year_events |>
  mutate(interrupted_flag = .data[[variants[[OUTCOME_VAR]]]])
```

# Part III — The model dataset

## 10. Assembling the activity-year panel

```{r}
model_data <- year_events |>
  select(iatiidentifier, year, interrupted_flag, balance_owed, balance_share,
         disbursement_usd, commitment_usd) |>
  # Activity-level attributes repeat across the activity's years.
  left_join(
    clean |> select(iatiidentifier, donor_ref, donor, x1_donor_type, modality,
                    recipient_main, status_code, status_name,
                    finance_type, flow_type, collaboration_type,
                    any_of(c("finance_type_name", "flow_type_name",
                             "collaboration_type_name")),
                    hierarchy, log_commitment,
                    planned_duration_days, y2_early_days, y1_gap_full),
    by = "iatiidentifier"
  ) |>
  # Structural position is already donor-country-year, so it joins on the
  # outcome's own unit rather than being averaged over an activity's life —
  # the averaging step the activity-level outcome used to require.
  # Built on commitments, so no lag is needed: a donor that withholds does not
  # thereby reduce what it has committed, and the measure is structurally prior
  # to the outcome. `disb_year` travels along only for
  # the descriptive comparison in 04_eda_clean.
  left_join(
    structural |> select(donor_ref, country, year, s1_share, s2_hhi_net,
                         s3_n_donors, cum_commit, disb_year),
    by = c("donor_ref", "recipient_main" = "country", "year")
  ) |>
  left_join(
    controls |> select(iso2, year, voice_account, corruption_ctrl, pol_stability,
                       aid_dependence, gdp_pc, population, region, income_group),
    by = c("recipient_main" = "iso2", "year")
  ) |>
  left_join(
    political |> select(iso2, year, coup, coup_attempt, election_year),
    by = c("recipient_main" = "iso2", "year")
  ) |>
  mutate(
    # S1 = 1 in a single-donor cell is maximal leverage, not an artefact
    #. Flagged rather than excluded; H3 drops these
    # because a pool net of its only member has no fragmentation to measure.
    single_donor = as.integer(s3_n_donors == 1),
    log_gdp_pc   = log(gdp_pc),
    log_pop      = log(population)
  )

model_data |>
  summarise(
    activity_years = n(),
    activities     = n_distinct(iatiidentifier),
    donors         = n_distinct(donor_ref),
    countries      = n_distinct(recipient_main),
    years          = paste(min(year), max(year), sep = "-"),
    interrupted_rate = round(100 * mean(interrupted_flag))
  )
```

## 11. Missingness in the assembled panel

The split between the two treatments declared earlier: what multiple imputation
will handle for H2 and H3, and what listwise deletion costs H1.

```{r}
model_data |>
  summarise(across(c(s1_share, s2_hhi_net, s3_n_donors, voice_account,
                     corruption_ctrl, pol_stability, aid_dependence,
                     log_gdp_pc, coup, election_year),
                   ~ round(100 * mean(is.na(.x))))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  arrange(desc(pct_missing))
```

```{r}
model_data |>
  summarise(
    complete_for_h1 = sum(complete.cases(pick(s1_share, voice_account,
                                              aid_dependence, log_gdp_pc))),
    total           = n()
  ) |>
  mutate(pct_retained = round(100 * complete_for_h1 / total))
```

## 12. Save

```{r}
write_csv(model_data, "../data/processed/model_data_year.csv")
write_csv(clean,      "../data/processed/activities_clean.csv")
```

**Output.** `data/processed/model_data_year.csv`, one row per activity-year,
with the refined outcome, structural position joined on its own unit, and both
control blocks attached. This is what 04_eda_clean describes and what the
models consume.
