---
title: "Network position"
subtitle: "The donor-recipient bipartite structure, and whether interruptions cluster"
author: "Gaspar Vigneaux"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
execute:
  warning: false
  message: false
---

S1, S2 and S3 describe a donor's position **inside one recipient's pool**. They
say nothing about where that donor sits in the system as a whole, and the
theory needs both: a donor present in forty countries faces a different
calculus when considering withdrawal from one of them than a donor present in
three.

This document adds the system-level view, and tests one hypothesis the
activity-year models cannot reach — whether interruptions cluster across donors
within a recipient more than chance allows.

**The edges are built on commitments, for the same reason S1 is**: a donor
that withholds a disbursement has not withdrawn its commitment, so a
commitment-based tie is not erased by the behaviour under study. A
disbursement-based network would delete precisely the edges of interest.

```{r}
#| label: setup
library(tidyverse)
library(igraph)
library(conflicted)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
conflict_prefer("simplify", "igraph")
conflict_prefer("union", "dplyr")

structural <- read_csv("../data/interim/structural.csv", show_col_types = FALSE)
model_data <- read_csv("../data/processed/model_data_year.csv", show_col_types = FALSE)
```

## 1. The bipartite structure

Two node classes, links only between them: donors on one side, recipients on
the other. A donor is tied to a recipient in a given year if it holds a live
cumulative commitment there.

```{r}
#| label: edges
edges <- structural |>
  filter(cum_commit > 0) |>
  transmute(donor = donor_ref, recipient = country, year,
            weight = cum_commit)

edges |>
  summarise(
    ties       = n(),
    donors     = n_distinct(donor),
    recipients = n_distinct(recipient),
    years      = n_distinct(year)
  )
```

```{r}
#| label: fig-density
#| fig-cap: "Ties, donors and recipients present in each year"
edges |>
  summarise(donors = n_distinct(donor), recipients = n_distinct(recipient),
            ties = n(), .by = year) |>
  pivot_longer(-year) |>
  ggplot(aes(year, value, colour = name)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  labs(x = NULL, y = NULL, colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## 2. Donor position, year by year

One graph per year, so position varies over the panel rather than being frozen
at a single snapshot. Centrality answers different questions depending on the
metric, and **choosing the metric is the hypothesis**:

| Question | Metric | What it means here |
|---|---|---|
| How many recipients? | Degree | Portfolio breadth — how diversified the donor's exposure is |
| Does it bridge? | Betweenness | Whether the donor links recipient groups that share no other funder |
| Tied to well-connected recipients? | Eigenvector | Whether the donor operates where the crowded pools are |

```{r}
#| label: donor-centrality
donor_year <- edges |>
  group_split(year) |>
  map(\(d) {
    g <- graph_from_data_frame(d |> select(donor, recipient), directed = FALSE)
    V(g)$type <- V(g)$name %in% d$recipient   # bipartite marker
    tibble(
      donor       = V(g)$name,
      is_donor    = !V(g)$type,
      year        = d$year[1],
      net_degree  = degree(g),
      net_between = betweenness(g, normalized = TRUE),
      net_eigen   = eigen_centrality(g)$vector
    ) |>
      filter(is_donor) |>
      select(-is_donor)
  }) |>
  list_rbind()

donor_year |>
  summarise(across(c(net_degree, net_between, net_eigen),
                   list(median = median, max = max))) |>
  pivot_longer(everything())
```

```{r}
#| label: fig-degree
#| fig-cap: "Portfolio breadth: recipients per donor, by year"
donor_year |>
  ggplot(aes(factor(year), net_degree)) +
  geom_boxplot(fill = "grey85", outlier.size = 0.8) +
  labs(x = NULL, y = "Recipients funded") +
  theme_minimal(base_size = 11)
```

- Degree is the measure the theory most directly needs. A donor funding many
  recipients spreads its reputational and relational exposure; withdrawing from
  one is a smaller share of its total engagement than for a donor with two or
  three partners. Neither S1 nor S2 sees this, because both are computed inside
  a single recipient.

## 3. The donor-donor projection, and what it does not mean

Projecting the bipartite graph onto its donor side ties two donors whenever
they fund the same recipient.

**A caution carried from the Marvel co-appearance project**: co-presence in a
recipient is *not* an observed relationship between donors. The projection is
an artefact of the data-generating process — two donors funding Malawi have not
thereby been shown to coordinate. What the measure captures is **exposure to the
same peers**, which is the observable footprint of the collective-action
structure the theory describes, not evidence of collaboration.

```{r}
#| label: projection
donor_peers <- edges |>
  group_split(year) |>
  map(\(d) {
    g <- graph_from_data_frame(d |> select(donor, recipient), directed = FALSE)
    V(g)$type <- V(g)$name %in% d$recipient
    proj <- bipartite_projection(g, which = "false")   # donor side
    tibble(
      donor          = V(proj)$name,
      year           = d$year[1],
      net_peers      = degree(proj),
      net_peer_str   = strength(proj)
    )
  }) |>
  list_rbind()

donor_peers |>
  summarise(median_peers = median(net_peers), max_peers = max(net_peers),
            .by = year) |>
  arrange(year)
```

## 4. The donor-year variable set

```{r}
#| label: assemble
network_vars <- donor_year |>
  left_join(donor_peers, by = c("donor", "year")) |>
  rename(donor_ref = donor)

network_vars |> glimpse()

write_csv(network_vars, "../data/processed/network_donor_year.csv")
```

These attach to the activity-year panel on `donor_ref` x `year`, alongside
S1-S3 rather than in place of them: the two sets answer different questions,
one about a donor's weight inside a recipient, the other about its position
across recipients.

```{r}
#| label: join-check
model_data |>
  left_join(network_vars, by = c("donor_ref", "year")) |>
  summarise(
    rows            = n(),
    net_degree_ok   = sum(!is.na(net_degree)),
    pct_matched     = round(100 * mean(!is.na(net_degree)))
  )
```

## 5. Do interruptions cluster within a recipient?

The collective-action account implies donors watch each other. If it holds,
interruptions inside a recipient-year should co-occur across donors more often
than independent decisions would produce.

**The null is constructed, not assumed.** Each donor's interruptions are
reshuffled across the years it was actually present, which preserves how often
each donor interrupts and how many donors each recipient has, while destroying
any within-recipient synchrony. Anything left is the effect.

```{r}
#| label: contagion
set.seed(20260824)   # fixed: a permutation test is stochastic

obs_panel <- model_data |>
  filter(!is.na(interrupted_flag)) |>
  distinct(donor_ref, recipient_main, year, interrupted_flag)

# Observed: mean number of donors interrupting in the same recipient-year,
# counting only recipient-years where more than one donor is present.
co_occurrence <- function(df) {
  df |>
    summarise(n_donors = n(), n_interrupted = sum(interrupted_flag),
              .by = c(recipient_main, year)) |>
    filter(n_donors > 1) |>
    summarise(mean_share = mean(n_interrupted / n_donors)) |>
    pull(mean_share)
}

observed <- co_occurrence(obs_panel)

null_dist <- map_dbl(1:1000, \(i) {
  obs_panel |>
    mutate(interrupted_flag = sample(interrupted_flag), .by = donor_ref) |>
    co_occurrence()
})

tibble(
  observed   = round(observed, 4),
  null_mean  = round(mean(null_dist), 4),
  null_sd    = round(sd(null_dist), 4),
  p_value    = round(mean(null_dist >= observed), 4)
)
```

```{r}
#| label: fig-contagion
#| fig-cap: "Observed within-recipient co-interruption against a permutation null"
tibble(null = null_dist) |>
  ggplot(aes(null)) +
  geom_histogram(bins = 40, fill = "grey70") +
  geom_vline(xintercept = observed, colour = "firebrick", linewidth = 0.9) +
  labs(x = "Mean share of a recipient's donors interrupting in the same year",
       y = "Permutations",
       caption = "Red line: observed value. 1,000 permutations, seed 20260824") +
  theme_minimal(base_size = 11)
```

**Reading the result.** A p-value near zero means interruptions synchronise
beyond what independent donor behaviour would produce — consistent with donors
responding to the same recipient-level shock, or to each other. The test cannot
distinguish those two: both produce clustering. Establishing which would need
the timing of the decisions rather than their annual co-occurrence, which the
data do not carry. A non-significant result is equally informative, and would
say that the collective-action account has no observable footprint here.

## 6. Save

```{r}
#| label: save
write_csv(network_vars, "../data/processed/network_donor_year.csv")
```

**Output.** `data/processed/network_donor_year.csv`, one row per donor-year,
carrying portfolio breadth, brokerage and peer exposure. 04_eda_clean
describes these alongside the rest of the variable set.
