---
title: "EDA II — The analytical sample"
subtitle: "What the models see, once the cleaning is settled"
author: "Gaspar Vigneaux"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    theme: cosmo
    fontsize: 1.02em
    linestretch: 1.5
    max-width: 1100px
    fig-width: 8
    fig-height: 4.5
    fig-align: center
    fig-cap-location: bottom
    tbl-cap-location: top
    df-print: kable
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

Three documents come before this one. 01_eda_raw diagnosed the data and
chose the outcome measure. 02_feature_engineering cleaned the variables.
03_network built the network measures. This document describes the result
and produces the figures for the data chapter.

**This document makes no decisions.** They were all made earlier, using criteria
set in advance. That order matters: choosing a measure
after seeing how it behaves is not a test.

**How to read the axes.** The unit of observation is the **activity-year**: one
operation in one year. Two things are easy to confuse, so they are labelled
differently.

- *"(count)"* means a number of activity-years.
- *"share of..."* means a percentage of some group's activity-years.

```{r}
#| label: setup
library(tidyverse)
library(scales)
library(patchwork)
library(kableExtra)
library(conflicted)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")

md <- read_csv("../data/processed/model_data_year.csv", show_col_types = FALSE)
nv <- read_csv("../data/processed/network_donor_year.csv", show_col_types = FALSE)

# One theme for every figure: a reader should not have to re-learn the visual
# conventions between panels.
theme_set(
  theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold", size = 11),
      plot.subtitle    = element_text(colour = "grey35", size = 9.5),
      plot.caption     = element_text(colour = "grey45", size = 8, hjust = 0),
      strip.text       = element_text(face = "bold", size = 9.5),
      legend.position  = "bottom"
    )
)

PAL_TYPE <- c(Bilateral = "#C0504D", Multilateral = "#4F6D8C")
INK      <- "grey25"

pc <- function(x) round(100 * mean(x, na.rm = TRUE))

# Tables are rendered rather than printed. A console tibble shows type headers,
# row numbers and truncated strings, none of which the reader needs.
show_tbl <- function(df, caption = NULL, full = FALSE) {
  df |>
    knitr::kable(caption = caption, format.args = list(big.mark = ","),
                 align = "l") |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = full, position = "left", font_size = 13
    )
}

d_all <- md |> left_join(nv, by = c("donor_ref", "year"))
```

# Part 0 — The disclosure filter

The outcome measures something that did **not** happen: a year where money was
owed and no payment arrived.

This only works if we would have seen the payment, had there been one. Some
donors publish their commitments to IATI but never publish their payments. For
those donors every year looks like a year with no payment, whatever they
actually did. The measure tells us nothing about them, so they are removed.

```{r}
#| label: tbl-disclosure
d_all |>
  summarise(`Activity-years` = n(),
            `Reports a payment` = round(100 * mean(disbursement_usd > 0)),
            `Flagged interrupted` = pc(interrupted_flag),
            .by = c(donor, x1_donor_type)) |>
  arrange(`Reports a payment`) |>
  rename(Donor = donor, Type = x1_donor_type) |>
  mutate(across(c(`Reports a payment`, `Flagged interrupted`),
                ~ paste0(.x, "%"))) |>
  show_tbl("Donors ordered by how often they publish a payment. The first six are removed.")
```

```{r}
#| label: apply-filter
# Donor-level, not activity-level. An activity with no disbursement under a
# donor that generally reports them is signal — possibly total withholding.
# Under a donor that never reports them it is an artefact.
MIN_DISCLOSURE <- 0.25

donors_keep <- d_all |>
  summarise(rate = mean(disbursement_usd > 0), .by = donor_ref) |>
  filter(rate >= MIN_DISCLOSURE) |>
  pull(donor_ref)

d <- d_all |> filter(donor_ref %in% donors_keep)
```

```{r}
#| label: fig-filter
#| fig-cap: "Donors ordered by how often they report a disbursement. The six below the line are excluded: for them a year without a disbursement is not evidence of anything, because they do not publish disbursements at all."
#| fig-height: 5.5
disc <- d_all |>
  summarise(n = n(),
            disclosing  = 100 * mean(disbursement_usd > 0),
            interrupted = 100 * mean(interrupted_flag),
            has_commit  = any(!is.na(balance_share)),
            .by = c(donor, x1_donor_type)) |>
  mutate(
    donor  = fct_reorder(str_trunc(donor, 34), disclosing),
    status = case_when(
      disclosing >= 25              ~ "kept",
      has_commit                    ~ "excluded: commitments but no disbursements",
      TRUE                          ~ "excluded: neither reported"
    )
  )

PAL_ST <- c("kept" = "grey35",
            "excluded: commitments but no disbursements" = "#C0504D",
            "excluded: neither reported" = "grey70")

p_disc <- disc |>
  ggplot(aes(disclosing, donor, colour = status)) +
  geom_segment(aes(x = 0, xend = disclosing, yend = donor), linewidth = 0.5) +
  geom_point(size = 2.6) +
  geom_vline(xintercept = 25, linetype = "dashed", colour = "grey45") +
  scale_colour_manual(values = PAL_ST) +
  scale_x_continuous(labels = label_percent(scale = 1), limits = c(0, 100)) +
  labs(x = "Years with a disbursement reported",
       y = NULL, colour = NULL,
       title = "How often each donor publishes a disbursement")

p_int <- disc |>
  ggplot(aes(interrupted, donor, colour = status)) +
  geom_segment(aes(x = 0, xend = interrupted, yend = donor), linewidth = 0.5) +
  geom_point(size = 2.6) +
  geom_text(aes(x = 100, label = paste0("n=", n)), hjust = 1, size = 2.6,
            colour = "grey50") +
  scale_colour_manual(values = PAL_ST, guide = "none") +
  scale_x_continuous(labels = label_percent(scale = 1), limits = c(0, 100)) +
  labs(x = "Years flagged as interrupted", y = NULL,
       title = "and how often it is flagged") +
  theme(axis.text.y = element_blank())

p_disc + p_int + plot_layout(guides = "collect", widths = c(1.35, 1))
```

**How to read it.** Each donor has one row. The row stays in the same place in
both panels, so only the horizontal position changes.

- **Red donors** publish commitments but almost never a payment. They are
  flagged in 94 to 100 per cent of their years. That number describes how they
  report, not what they do.
- **Grey donors** publish neither. They never owe anything, so they are never
  flagged. They are removed too, because they add nothing either way.

**Check the `n` labels.** New Zealand and Denmark have one activity-year each.
Their rates of 100 and 0 per cent come from a single observation, not a pattern.
Sample size is written as text, not as point size: donors range from 1 to 857
observations, and size would make the small ones disappear.

```{r}
#| label: tbl-filter-effect
bind_rows(
  d_all |> summarise(Sample = "Before filter", rate = pc(interrupted_flag),
                     n = n(), .by = x1_donor_type),
  d     |> summarise(Sample = "After filter",  rate = pc(interrupted_flag),
                     n = n(), .by = x1_donor_type)
) |>
  transmute(Type = x1_donor_type, Sample,
            `Activity-years` = n, `Interrupted` = paste0(rate, "%")) |>
  arrange(Type, desc(Sample)) |>
  show_tbl("What the filter changes")
```

- **The gap between donor types was about twice as large before the filter.**
  Two publishers caused that on their own.
- This belongs in the results chapter, not only in the limitations. Without the
  filter, a comparison between bilateral and multilateral donors partly measures
  how they report rather than how they behave.

# Part 1 — The sample

```{r}
#| label: tbl-sample
d |>
  summarise(`Activity-years` = n(),
            Activities = n_distinct(iatiidentifier),
            Donors = n_distinct(donor_ref),
            Countries = n_distinct(recipient_main),
            Window = paste(min(year), max(year), sep = "-"),
            Interrupted = paste0(pc(interrupted_flag), "%")) |>
  show_tbl("The analytical sample")

d |>
  summarise(`Activity-years` = n(),
            Activities = n_distinct(iatiidentifier),
            Donors = n_distinct(donor_ref),
            Countries = n_distinct(recipient_main),
            Interrupted = paste0(pc(interrupted_flag), "%"),
            .by = x1_donor_type) |>
  rename(Type = x1_donor_type) |>
  show_tbl("By institutional type")
```

- **`r pc(d$x1_donor_type == "Multilateral")` per cent of activity-years are
  multilateral.** The research question compares the two donor types, and one
  type holds four fifths of the data. No modelling choice fixes this. It limits
  how precise the comparison can be.

## The donor level

```{r}
#| label: fig-donors
#| fig-cap: "Observations per donor and their interruption rate. The horizontal scale is logarithmic because on a linear axis the five largest donors compress the remaining twelve into an unreadable band."
#| fig-height: 5.5
donor_tab <- d |>
  summarise(activity_years = n(),
            countries = n_distinct(recipient_main),
            interrupted = pc(interrupted_flag),
            .by = c(donor, x1_donor_type)) |>
  mutate(donor = str_trunc(donor, 38)) |>
  arrange(desc(activity_years))

donor_tab |>
  transmute(Donor = donor, Type = x1_donor_type,
            `Activity-years` = activity_years, Countries = countries,
            Interrupted = paste0(interrupted, "%")) |>
  show_tbl("Every donor in the analytical sample")

p_n <- donor_tab |>
  mutate(donor = fct_reorder(donor, activity_years)) |>
  ggplot(aes(activity_years, donor, fill = x1_donor_type)) +
  geom_col(width = 0.72) +
  scale_fill_manual(values = PAL_TYPE) +
  scale_x_log10() +
  labs(x = "Activity-years observed (count, log scale)", y = NULL, fill = NULL,
       title = "Sample weight")

p_r <- donor_tab |>
  mutate(donor = fct_reorder(donor, activity_years)) |>
  ggplot(aes(interrupted, donor, colour = x1_donor_type)) +
  geom_segment(aes(x = 0, xend = interrupted, yend = donor), linewidth = 0.5) +
  geom_point(size = 2.4) +
  scale_colour_manual(values = PAL_TYPE) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(x = "Share of the donor's years interrupted", y = NULL, colour = NULL,
       title = "Interruption rate") +
  theme(axis.text.y = element_blank())

p_n + p_r + plot_layout(guides = "collect", widths = c(1.4, 1))
```

- **Donors range from one observation to several hundred.** This is why the
  model uses random donor effects rather than fixed effects. With fixed
  effects, a donor with one observation is fitted perfectly and tells us
  nothing. Random effects weight each donor by how much data it actually has.

# Part 2 — The outcome

```{r}
#| label: fig-outcome
#| fig-cap: "Interruption over time, overall and by institutional type. Point size carries the number of activity-years behind each rate, so that a rate computed on twenty observations is not read as equivalent to one computed on four hundred."
#| fig-height: 4
p_year <- d |>
  summarise(rate = 100 * mean(interrupted_flag), n = n(), .by = year) |>
  ggplot(aes(year, rate)) +
  geom_col(fill = INK, width = 0.68) +
  scale_x_continuous(breaks = 2014:2023) +
  scale_y_continuous(labels = label_percent(scale = 1),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(x = NULL, y = "Share of activity-years interrupted", title = "All donors")

p_type <- d |>
  summarise(rate = 100 * mean(interrupted_flag), n = n(),
            .by = c(year, x1_donor_type)) |>
  ggplot(aes(year, rate, colour = x1_donor_type)) +
  geom_line(linewidth = 0.6) +
  geom_point(aes(size = n)) +
  scale_colour_manual(values = PAL_TYPE) +
  scale_size_continuous(range = c(1, 4.5), guide = "none") +
  scale_x_continuous(breaks = 2014:2023) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(x = NULL, y = NULL, colour = NULL, title = "By institutional type")

p_year + p_type + plot_layout(widths = c(1, 1.2))
```

# Part 3 — Structural position

## What these three variables mean

Before the shapes, what the numbers are. All three describe one donor in one
country in one year, and all three are built from **commitments** — money
promised — rather than from money actually paid.

Take Malawi in 2019. Three donors have promised budget support, and these are
their running totals:

| Donor | Promised so far | Share of the total |
|---|---|---|
| World Bank | 200 M | **53%** |
| Germany | 100 M | 26% |
| European Union | 80 M | 21% |
| **Total promised to Malawi** | **380 M** | 100% |

For the World Bank's row in 2019:

- **S1 = 0.53.** Just over half of everything promised to Malawi comes from the
  World Bank. This is what the hypothesis calls leverage: if the World Bank
  stops paying, Malawi cannot easily replace that money.
- **S2 = 0.11.** How concentrated the *other* donors are among themselves,
  ignoring the World Bank. Germany and the EU are similar in size, so the rest
  of the pool is spread out rather than dominated by one player.
- **S3 = 3.** Three donors are present.

Germany's row in the same country-year would read S1 = 0.26, and its S2 would
describe how the World Bank and the EU are distributed. Every donor sees a
different picture of the same country.

**Reading the bands later in this document:** a `<25%` band means the donor
supplies less than a quarter of what the country has been promised; a `>75%`
band means it supplies almost all of it.

## Are these distributions a problem?

They are not bell-shaped, and that is fine. Here is why.

**Normality is an assumption of the linear model.** It applies to the
*residuals*, not to the predictors. And the outcome here is binary — interrupted
or not — so the model is a logistic GLM. That model exists precisely because
real data break the linear model's assumptions. **No variable below needs to
follow any particular distribution.**

The shapes still matter for interpretation. Two are worth noting.

```{r}
#| label: fig-structural
#| fig-cap: "The structural variables. s1_share is a proportion with a deliberate mass at 1; s3_n_donors is a small count; neither is meant to be bell-shaped."
#| fig-height: 3.6
lbl <- c(s1_share = "S1 — the donor's share of all money\npromised to that country",
         s2_hhi_net = "S2 — how concentrated the\nother donors are",
         s3_n_donors = "S3 — how many donors\nthe country has")

d |>
  select(s1_share, s2_hhi_net, s3_n_donors) |>
  pivot_longer(everything(), names_to = "v", values_to = "value") |>
  filter(!is.na(value)) |>
  mutate(v = factor(lbl[v], levels = unname(lbl))) |>
  ggplot(aes(value)) +
  geom_histogram(bins = 30, fill = INK) +
  facet_wrap(~v, scales = "free", nrow = 1) +
  labs(x = NULL, y = "Activity-years (count)")
```

- **The spike at 1 in S1 is real, not a problem.** These are the
  `r pc(d$single_donor == 1)` per cent of years where the donor is the country's
  only budget support provider, so its share is 1 by definition. This is the
  strongest case for H2: maximum leverage. They are kept on purpose and marked
  with `single_donor`.
- **S3 counts donors and runs from 1 to 8.** A count is not supposed to look
  normal. But this range is narrow. The fragmentation literature studies
  countries with a dozen donors or more, and our data cannot reach that range.
  The write-up says so.

## The collinearity that must be handled before estimation

```{r}
#| label: fig-corr
#| fig-cap: "Correlation between the explanatory variables. The -0.87 between S1 and S2 is mechanical: S2 is calculated by subtracting S1 squared."
#| fig-height: 4.4
cm <- d |>
  select(`S1 donor share` = s1_share, `S2 others concentration` = s2_hhi_net,
         `S3 donors in country` = s3_n_donors, `Network degree` = net_degree,
         `log commitment` = log_commitment) |>
  cor(use = "pairwise.complete.obs")

cm |>
  as_tibble(rownames = "a") |>
  pivot_longer(-a, names_to = "b", values_to = "r") |>
  mutate(across(c(a, b), ~ factor(.x, levels = rownames(cm)))) |>
  ggplot(aes(a, b, fill = r)) +
  geom_tile(colour = "white", linewidth = 1) +
  geom_text(aes(label = number(r, accuracy = 0.01)), size = 3.1,
            colour = if_else(abs(as.vector(t(cm))) > 0.6, "white", "grey20")) +
  scale_fill_gradient2(low = "#C0504D", mid = "white", high = "#4F6D8C",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = NULL) +
  coord_fixed() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

- **S1 and S2 correlate at about -0.87, and this is by construction.** S2 is
  calculated as `s2_hhi - s1_share^2`, so it literally subtracts S1. Put both in
  one model and the standard errors inflate: both can look non-significant even
  when both matter. The models handle this with principal components, reported
  next to the single-variable models rather than replacing them.
- **Network degree correlates weakly with all three.** This is why it was worth
  building. It measures something the within-country variables do not.

# Part 4 — Donor size, as a control

S1, S2 and S3 describe a donor **inside one country**. This section adds one
variable describing the donor **overall**: how many countries it funds.

**It is a control, not a hypothesis.** The theory has two mechanisms,
substitutability and signalling, and neither says anything about a donor's total
size. The variable is here for a narrower reason: **multilaterals are usually
the bigger donors**, so institutional type and size are entangled. A result
reading "multilaterals behave differently" may really be "big donors behave
differently", with the institutional label doing no work of its own. Separating
them requires measuring both.

**The variable.** `net_degree` counts how many countries a donor funds in a
given year. It describes a donor in a year, not a single activity, so all of
that donor's activities in that year share the same value.

```{r}
#| label: fig-network
#| fig-cap: "Portfolio breadth by institutional type. Each point is one donor in one year."
#| fig-height: 4
nd <- d |> distinct(donor_ref, donor, year, x1_donor_type, net_degree) |>
  filter(!is.na(net_degree))

p_box <- nd |>
  ggplot(aes(x1_donor_type, net_degree, colour = x1_donor_type)) +
  geom_boxplot(outlier.shape = NA, width = 0.45, linewidth = 0.5) +
  geom_jitter(width = 0.13, alpha = 0.35, size = 1.3) +
  scale_colour_manual(values = PAL_TYPE, guide = "none") +
  labs(x = NULL, y = "Recipient countries the donor funds that year",
       title = "Multilaterals fund more countries")

p_hist <- nd |>
  ggplot(aes(net_degree, fill = x1_donor_type)) +
  geom_histogram(bins = 22, alpha = 0.75, position = "identity") +
  scale_fill_manual(values = PAL_TYPE) +
  labs(x = "Recipient countries the donor funds that year",
       y = "Donor-years (count)", fill = NULL,
       title = "but the two groups overlap")

p_box + p_hist
```

- **Multilaterals do fund more countries on average**, but the groups overlap:
  some multilaterals fund only a handful, some bilaterals fund many. The two
  variables are related without being the same thing, which is what makes the
  separation possible.

**What it buys for H1.** Put institutional type and donor size in the same
model, and see which one survives.

| If the surviving variable is... | Then the conclusion is |
|---|---|
| Institutional type | Something about how these organisations are governed matters |
| Donor size | The bilateral/multilateral label was standing in for size |
| Both | Each contributes something the other does not |

# Part 5 — Controls

```{r}
#| label: fig-controls
#| fig-cap: "Recipient-side controls. Aid dependence is right-skewed and enters logged; the governance indices are already standardised by their source."
#| fig-height: 5
ctrl_lbl <- c(voice_account = "Voice & accountability",
              corruption_ctrl = "Control of corruption",
              pol_stability = "Political stability",
              aid_dependence = "Aid dependence (ODA/GNI, %)",
              log_aid_dep = "Aid dependence (log)",
              log_gdp_pc = "GDP per capita (log)")

d |>
  # if_else evaluates both branches, so log() must not see the non-positive
  # values even though they are discarded. Negative ODA is net repayment.
  mutate(log_aid_dep = if_else(aid_dependence > 0,
                               log(pmax(aid_dependence, 1e-6)), NA_real_)) |>
  select(all_of(names(ctrl_lbl))) |>
  pivot_longer(everything(), names_to = "v", values_to = "value") |>
  filter(!is.na(value)) |>
  mutate(v = factor(ctrl_lbl[v], levels = unname(ctrl_lbl))) |>
  ggplot(aes(value)) +
  geom_histogram(bins = 30, fill = INK) +
  facet_wrap(~v, scales = "free", nrow = 2) +
  labs(x = NULL, y = "Activity-years (count)")
```

- **Aid dependence is heavily right-skewed.** Most countries receive modest aid
  relative to the size of their economy; a few receive very large amounts.
- Both versions are shown. The raw one makes the skew visible; the logged one
  goes into the models. Without the log, a handful of extreme countries would
  pull the estimates much harder than their numbers justify. This is about
  influence, not about any normality requirement.

```{r}
#| label: tbl-political
d |>
  summarise(`Coup years` = sum(coup, na.rm = TRUE),
            `Election years` = sum(election_year, na.rm = TRUE),
            `Share with a coup` = paste0(pc(coup), "%"),
            `Share with an election` = paste0(pc(election_year), "%")) |>
  show_tbl("Political events in the sample")
```

- **Coups are rare**, so that indicator will always be sparse. **Elections cover
  about a quarter of activity-years**, which is enough variation to be useful.
- Both describe the recipient country, so they apply equally to bilateral and
  multilateral donors. That is why they work as controls where donor-recipient
  voting affinity did not.

# Part 6 — First look at the hypotheses

These are raw patterns. **Nothing here controls for anything.** They exist so
the controlled estimates can be compared against them later.

## What each mechanism predicts

The two hypotheses use different variables, and the same mechanism can appear
with opposite signs depending on which one. Both S1 and S3 measure the **same
underlying thing** — how strong the donor's position is — but from opposite
directions:

| | Strong position | Weak position |
|---|---|---|
| **S1 — donor share** | High: the donor dominates | Low |
| **S3 — donors in the country** | **Few**: little competition | Many |

A high share means **few** competitors, not many. If one donor supplies 80 per
cent of what a country receives, there is hardly anyone else there.

**Substitutability** asks *"can I be replaced?"* A donor in a strong position
cannot, so its threat is credible and it uses it:

- more share → more interruption
- **fewer** donors → more interruption

One prediction, two expressions.

**Signalling** asks *"who is watching?"* The audience is the other donors, so:

- **more** donors → more interruption

The two mechanisms therefore disagree **only about S3**. Signalling says nothing
about share, because its logic concerns audience rather than weight. That is why
H2 is directional and H3 is an adjudication between rivals.

```{r}
#| label: fig-hypotheses
#| fig-cap: "Interruption against donor share and against pool size. Bands are ordered along the horizontal axis because the underlying variables are ordered — an unordered display of ordered data would misrepresent it."
#| fig-height: 4
h2 <- d |>
  filter(!is.na(s1_share)) |>
  mutate(band = cut(s1_share, c(0, .25, .5, .75, 1),
                    labels = c("<25%", "25-50%", "50-75%", ">75%"),
                    include.lowest = TRUE, ordered_result = TRUE)) |>
  summarise(rate = 100 * mean(interrupted_flag), n = n(), .by = band)

h3 <- d |>
  filter(!is.na(s3_n_donors)) |>
  mutate(band = cut(s3_n_donors, c(0, 1, 2, 3, 20),
                    labels = c("1", "2", "3", "4+"), ordered_result = TRUE)) |>
  summarise(rate = 100 * mean(interrupted_flag), n = n(), .by = band)

bar <- function(df, ttl, xlab) {
  df |>
    arrange(band) |>
    ggplot(aes(band, rate)) +
    geom_col(fill = INK, width = 0.62) +
    geom_text(aes(label = paste0("n=", n)), vjust = -0.45, size = 2.8,
              colour = "grey40") +
    scale_y_continuous(labels = label_percent(scale = 1),
                       expand = expansion(mult = c(0, 0.16))) +
    labs(x = xlab, y = "Share of activity-years interrupted", title = ttl)
}

bar(h2, "H2 — donor share",
    "The donor's share of all money promised to that country") +
  bar(h3, "H3 — pool size",
      "Number of budget support donors the country has")
```

```{r}
#| label: tbl-hypotheses
bind_rows(
  h2 |> arrange(band) |> transmute(Hypothesis = "H2 - donor share",
                                   Band = as.character(band),
                                   `Activity-years` = n,
                                   Interrupted = paste0(round(rate), "%")),
  h3 |> arrange(band) |> transmute(Hypothesis = "H3 - pool size",
                                   Band = as.character(band),
                                   `Activity-years` = n,
                                   Interrupted = paste0(round(rate), "%"))
) |>
  show_tbl("Raw interruption rates by band")
```

**Reading the two panels against the table above:**

- **H2.** The interruption rate rises with the donor's share. That is what
  substitutability predicts.
- **H3.** The interruption rate falls as the country gains donors. That is
  substitutability again — not signalling, which predicts the opposite.

**Both panels point to the same mechanism.** The signs are opposite because the
variables are opposite, not because the story is.

This also runs against Molenaers et al. (2015), who find that more donors
produce *more* suspensions. H3 was written as an adjudication between two rival
accounts, and the raw pattern adjudicates for substitutability, against the
published finding.

- **But these are descriptions, not tests.** Countries differ in governance and
  aid dependence. Donors differ in size. Both change systematically along the
  horizontal axis, and neither is held constant here. The models exist to check
  whether the pattern survives.

# Save

```{r}
#| label: save
d |> write_csv("../data/processed/model_data_with_network.csv")
```

**Output.** `data/processed/model_data_with_network.csv` — the activity-year
panel, disclosure-filtered, with network position attached. This is what
05_models and 06_prediction estimate on.
