---
title: "Models — H2 and H3"
subtitle: "Multilevel logistic regression, built one level at a time"
author: "Gaspar Vigneaux"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    theme: cosmo
    fontsize: 1.02em
    linestretch: 1.5
    max-width: 1100px
    fig-width: 8
    fig-height: 4.5
    fig-align: center
    embed-resources: true
execute:
  warning: false
  message: false
---

This document estimates **H2** and **H3**: what the mechanism is, and in which
direction it runs.

**H1 is not here.** It asks whether structural position *predicts better* than
institutional type — a comparison of out-of-sample performance, not of
coefficients. That is 06_prediction.

The model is built in steps rather than presented finished. Each step answers a
question the next one depends on: is there group-level variation at all, how
much, does it need random effects, and only then, what predicts the outcome.

```{r}
#| label: setup
library(tidyverse)
library(lme4)
library(broom.mixed)   # supplies the tidy() method pool() needs for glmerMod
library(mice)
library(kableExtra)
library(scales)
library(patchwork)
library(conflicted)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")

d <- read_csv("../data/processed/model_data_with_network.csv", show_col_types = FALSE)

theme_set(
  theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold", size = 11),
          plot.subtitle = element_text(colour = "grey35", size = 9.5),
          legend.position = "bottom")
)

PAL  <- c(Bilateral = "#C0504D", Multilateral = "#4F6D8C")
CTRL <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))

show_tbl <- function(df, caption = NULL) {
  df |>
    knitr::kable(caption = caption, digits = 3, align = "l") |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE, position = "left", font_size = 13)
}

tidy_pooled <- function(pooled, keep = NULL) {
  s <- summary(pooled, conf.int = TRUE) |> as_tibble()
  names(s)[names(s) == "2.5 %"]  <- "lo"
  names(s)[names(s) == "97.5 %"] <- "hi"
  s <- s |> filter(term != "(Intercept)")
  if (!is.null(keep)) s <- s |> filter(term %in% keep)
  s |>
    transmute(Term = term,
              `Odds ratio` = round(exp(estimate), 3),
              `95% CI` = paste0("[", round(exp(lo), 2), ", ", round(exp(hi), 2), "]"),
              p = round(p.value, 4))
}
```

```{r}
#| label: prepare
model_df <- d |>
  mutate(
    log_aid_dep = if_else(aid_dependence > 0, log(pmax(aid_dependence, 1e-6)), NA_real_),
    donor_type  = factor(x1_donor_type, levels = c("Bilateral", "Multilateral")),
    year_c      = year - 2014,
    # Standardised so coefficients are comparable in magnitude and the
    # optimiser converges more reliably.
    across(c(s1_share, s2_hhi_net, s3_n_donors, net_degree, voice_account,
             corruption_ctrl, log_aid_dep, log_gdp_pc, log_commitment),
           ~ as.numeric(scale(.x)), .names = "z_{.col}")
  )

MODEL_VARS <- c("interrupted_flag", "z_s1_share", "z_s2_hhi_net", "z_s3_n_donors",
                "z_net_degree", "z_voice_account", "z_corruption_ctrl",
                "z_log_aid_dep", "z_log_gdp_pc", "z_log_commitment",
                "coup", "election_year", "donor_type", "donor_ref",
                "recipient_main", "year_c")

cc <- model_df |> drop_na(all_of(MODEL_VARS))
```

**Two estimation samples, and why.** Likelihood-ratio tests and variance
components have no standard pooling rule across imputed datasets — a likelihood
ratio is not a quantity Rubin's rules combine. Model **building and
diagnostics** therefore run on complete cases; the **final estimates** are
pooled across imputations. Both samples are reported so the difference is
visible.

```{r}
#| label: samples
tibble(
  Sample = c("Complete cases (building, diagnostics)", "All rows (final estimates, imputed)"),
  `Activity-years` = c(nrow(cc), nrow(model_df)),
  Donors = c(n_distinct(cc$donor_ref), n_distinct(model_df$donor_ref)),
  Countries = c(n_distinct(cc$recipient_main), n_distinct(model_df$recipient_main))
) |>
  show_tbl("The two samples")
```

# Does the data need a multilevel model?

Not assumed — tested. The question is whether interruption clusters by donor and
by country beyond what chance produces.

## M0 and M1: the null models

```{r}
#| label: null-models
#| cache: true
m0 <- glm(interrupted_flag ~ 1, data = cc, family = binomial)

m1 <- glmer(interrupted_flag ~ 1 + (1 | donor_ref) + (1 | recipient_main),
            data = cc, family = binomial, control = CTRL)

# Computed directly rather than through anova(): anova.merMod compares the two
# models' data slots by identity, and a glm() and a glmer() never share one.
# The arithmetic is the same and it is visible here.
ll0 <- as.numeric(logLik(m0)); ll1 <- as.numeric(logLik(m1))
lrt <- 2 * (ll1 - ll0)
df_lrt <- 2                       # two variance components added

tibble(
  Model = c("M0 - single level", "M1 - random intercepts"),
  Parameters = c(attr(logLik(m0), "df"), attr(logLik(m1), "df")),
  logLik = round(c(ll0, ll1), 1),
  AIC = round(c(AIC(m0), AIC(m1))),
  BIC = round(c(BIC(m0), BIC(m1)))
) |>
  show_tbl("Model fit with and without the grouping structure")

tibble(
  `LRT statistic` = round(lrt, 1),
  df = df_lrt,
  `p (naive)` = format.pval(pchisq(lrt, df_lrt, lower.tail = FALSE),
                            digits = 3, eps = 1e-16)
) |>
  show_tbl("Likelihood-ratio test: is the grouping structure needed?")
```

- The test compares a model with no grouping against one with random intercepts
  for donor and country. A significant result means observations inside the same
  donor, or the same country, are more alike than a single-level model allows.
- **The p-value is labelled naive on purpose.** Testing whether a variance is
  zero puts the null on the **boundary** of the parameter space, because a
  variance cannot be negative. The usual chi-squared reference distribution is
  then wrong in a known direction: it is too conservative. The true p-value is
  smaller than the one reported, so a rejection here is a safe rejection.

## The variance partition coefficient

For a binary outcome fitted with a logit link, the level-one variance is not
estimated. It is fixed by the logistic distribution at **π²/3 ≈ 3.29**, and the
VPC is computed against that constant.

```{r}
#| label: vpc
vc <- as_tibble(VarCorr(m1))
L1 <- pi^2 / 3

vc |>
  transmute(Level = if_else(grp == "donor_ref", "Donor", "Recipient country"),
            Variance = round(vcov, 3),
            SD = round(sdcor, 3),
            VPC = round(vcov / (sum(vc$vcov) + L1), 3)) |>
  bind_rows(tibble(Level = "Activity-year (fixed at pi^2/3)",
                   Variance = round(L1, 3), SD = round(sqrt(L1), 3),
                   VPC = round(L1 / (sum(vc$vcov) + L1), 3))) |>
  show_tbl("Where the variation sits, before any predictor is added")
```

```{r}
#| label: vpc-values
#| echo: false
vpc_d <- vc$vcov[vc$grp == "donor_ref"]   / (sum(vc$vcov) + L1)
vpc_c <- vc$vcov[vc$grp == "recipient_main"] / (sum(vc$vcov) + L1)
```

- **`r round(100 * vpc_d)` per cent of the variation sits between donors** and
  **`r round(100 * vpc_c)` per cent between countries**, before any predictor
  enters. The rest is within.
- That the donor share is the larger of the two is itself substantive: *who* is
  paying explains more of whether a payment stops than *where* the money goes.

# How much do donors differ?

The random-intercept model assumes donors differ in their baseline rate. This
section asks whether they also differ in **how they respond to leverage**, by
fitting one model per donor and comparing.

Only donors with enough observations can support a separate fit, so this is
restricted rather than complete — which is itself an argument for pooling.

```{r}
#| label: per-donor
#| cache: true
MIN_N <- 100

per_donor <- cc |>
  add_count(donor_ref, name = "n_donor") |>
  filter(n_donor >= MIN_N) |>
  nest(.by = c(donor, donor_type)) |>
  mutate(
    fit = map(data, \(x) glm(interrupted_flag ~ z_s1_share + z_log_aid_dep +
                               z_log_commitment, data = x, family = binomial)),
    tidied = map(fit, \(f) broom::tidy(f, conf.int = TRUE))
  ) |>
  select(donor, donor_type, tidied) |>
  unnest(tidied) |>
  filter(term == "z_s1_share")

per_donor |>
  transmute(Donor = str_trunc(donor, 34), Type = donor_type,
            `Odds ratio` = round(exp(estimate), 2),
            `95% CI` = paste0("[", round(exp(conf.low), 2), ", ",
                              round(exp(conf.high), 2), "]")) |>
  show_tbl("The share effect, estimated separately for each large donor")
```

```{r}
#| label: fig-per-donor
#| fig-cap: "The effect of donor share on interruption, fitted separately per donor. Wide intervals and differing signs are what partial pooling exists to handle."
#| fig-height: 3.6
per_donor |>
  mutate(donor = fct_reorder(str_trunc(donor, 30), estimate)) |>
  ggplot(aes(exp(estimate), donor, colour = donor_type)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = exp(conf.low), xmax = exp(conf.high)), size = 0.45) +
  scale_colour_manual(values = PAL) +
  scale_x_log10() +
  labs(x = "Odds ratio for donor share (log scale)", y = NULL, colour = NULL,
       title = "Donors do not all respond to leverage the same way")
```

# Building the model up

Predictors are added one level at a time, so it is visible what each level
contributes rather than only what the final model reports.

```{r}
#| label: build
#| cache: true
f2 <- interrupted_flag ~ z_s1_share + z_log_commitment + coup + election_year +
  year_c + (1 | donor_ref) + (1 | recipient_main)

f3 <- update(f2, . ~ . + z_voice_account + z_corruption_ctrl + z_log_aid_dep +
               z_log_gdp_pc)

f4 <- update(f3, . ~ . + donor_type + z_net_degree)

m2 <- glmer(f2, data = cc, family = binomial, control = CTRL)
m3 <- glmer(f3, data = cc, family = binomial, control = CTRL)
m4 <- glmer(f4, data = cc, family = binomial, control = CTRL)

tibble(
  Model = c("M1 - null", "M2 - + activity-year", "M3 - + country controls",
            "M4 - + donor level"),
  Adds = c("random intercepts only", "share, size, coup, election, year",
           "governance, aid dependence, income", "institutional type, donor size"),
  AIC = round(c(AIC(m1), AIC(m2), AIC(m3), AIC(m4))),
  BIC = round(c(BIC(m1), BIC(m2), BIC(m3), BIC(m4))),
  `SD donor` = round(c(as_tibble(VarCorr(m1))$sdcor[as_tibble(VarCorr(m1))$grp == "donor_ref"],
                       as_tibble(VarCorr(m2))$sdcor[as_tibble(VarCorr(m2))$grp == "donor_ref"],
                       as_tibble(VarCorr(m3))$sdcor[as_tibble(VarCorr(m3))$grp == "donor_ref"],
                       as_tibble(VarCorr(m4))$sdcor[as_tibble(VarCorr(m4))$grp == "donor_ref"]), 3)
) |>
  show_tbl("Model comparison as predictors are added")
```

- Watch the donor standard deviation across rows. If it falls when donor-level
  predictors enter at M4, those predictors are explaining part of what the
  random effect was absorbing. If it does not, the differences between donors
  are something these variables do not capture.

## The same build-up, coefficient by coefficient

The table above shows what each block does to fit. This one shows what it does
to every estimate: the same four models, one column each, coefficients reported
as odds ratios with their 95 per cent interval underneath — the scale the rest
of this document interprets in, rather than the raw log-odds a `summary()` call
would print.

```{r}
#| label: regression-table
models <- list(`(1)` = m1, `(2)` = m2, `(3)` = m3, `(4)` = m4)
stars  <- function(p) case_when(p < 0.01 ~ "***", p < 0.05 ~ "**", p < 0.1 ~ "*", TRUE ~ "")

# The order and labelling blocks match f2/f3/f4 above; the intercept is
# omitted, since with two centred random effects it is not a quantity anyone
# interprets on its own.
TERM_ORDER <- c("z_s1_share", "z_log_commitment", "coup", "election_year", "year_c",
                "z_voice_account", "z_corruption_ctrl", "z_log_aid_dep", "z_log_gdp_pc",
                "donor_typeMultilateral", "z_net_degree")
TERM_LBL <- c(z_s1_share = "Donor share", z_log_commitment = "Operation size (log)",
              coup = "Coup", election_year = "Election year", year_c = "Year",
              z_voice_account = "Voice and accountability",
              z_corruption_ctrl = "Control of corruption",
              z_log_aid_dep = "Aid dependence (log)", z_log_gdp_pc = "GDP per capita (log)",
              donor_typeMultilateral = "Multilateral (ref: bilateral)",
              z_net_degree = "Donor portfolio breadth")

coef_tbl <- imap_dfr(models, \(m, lbl) {
  broom.mixed::tidy(m, effects = "fixed", conf.int = TRUE) |>
    filter(term != "(Intercept)") |>
    transmute(term, model = lbl,
              cell = paste0(round(exp(estimate), 3), stars(p.value), "<br>(",
                            round(exp(conf.low), 2), ", ", round(exp(conf.high), 2), ")"))
}) |>
  pivot_wider(names_from = model, values_from = cell)

# M1 has no fixed effects besides the intercept, which is dropped above, so its
# column never appears in the pivot. Added back empty rather than assumed
# present, since a `(1)`:`(4)` range selector below would otherwise error.
coef_tbl[setdiff(names(models), names(coef_tbl))] <- NA_character_

coef_tbl <- coef_tbl |>
  mutate(term = factor(term, levels = TERM_ORDER)) |>
  arrange(term) |>
  mutate(term = recode(as.character(term), !!!TERM_LBL)) |>
  select(term, all_of(names(models))) |>
  mutate(across(all_of(names(models)), \(x) replace_na(x, "")))

vc_tbl <- imap_dfr(models, \(m, lbl) {
  vc <- as_tibble(VarCorr(m))
  tibble(term = c("SD (donor)", "SD (recipient country)"), model = lbl,
         cell = as.character(round(c(vc$sdcor[vc$grp == "donor_ref"],
                                     vc$sdcor[vc$grp == "recipient_main"]), 3)))
}) |>
  pivot_wider(names_from = model, values_from = cell)

fit_tbl <- imap_dfr(models, \(m, lbl) {
  tibble(term = c("Activity-years", "Donors", "Recipient countries",
                  "Log likelihood", "AIC", "BIC"),
         model = lbl,
         cell = c(nobs(m), n_distinct(cc$donor_ref), n_distinct(cc$recipient_main),
                  round(as.numeric(logLik(m)), 1), round(AIC(m)), round(BIC(m))) |>
           as.character())
}) |>
  pivot_wider(names_from = model, values_from = cell)

bind_rows(coef_tbl, vc_tbl, fit_tbl) |>
  knitr::kable(caption = "Interruption, built up one block at a time (odds ratios; 95% CI in parentheses)",
               align = c("l", "c", "c", "c", "c"), escape = FALSE) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE, position = "left", font_size = 13) |>
  kableExtra::add_header_above(c(" " = 1, "Probability of interruption" = 4)) |>
  kableExtra::pack_rows("Variance components", nrow(coef_tbl) + 1,
                        nrow(coef_tbl) + nrow(vc_tbl)) |>
  kableExtra::pack_rows("Model fit", nrow(coef_tbl) + nrow(vc_tbl) + 1,
                        nrow(coef_tbl) + nrow(vc_tbl) + nrow(fit_tbl)) |>
  kableExtra::footnote(
    general_title = "Note:",
    general = "Complete cases (n = 1,836 activity-years, 15 donors, 94 recipient countries). *p<0.1; **p<0.05; ***p<0.01.")
```

- Read this table and the pooled H2/H3 tables below as different things. This
  one is fit on complete cases and exists to show what each block of predictors
  does to the estimates as it enters — it is a diagnostic on the model, not the
  final result. The pooled tables that follow are the estimates that are
  actually interpreted, combined across the five imputed datasets by Rubin's
  rules. Complete-case and pooled coefficients are close throughout, which is
  itself part of what supports H2 in §4.

## Does the effect of leverage vary by donor?

The per-donor fits above suggest it might. A random slope tests it inside one
model, borrowing strength across donors instead of fitting each alone.

```{r}
#| label: random-slope
#| cache: true
m5 <- try(
  glmer(update(f4, . ~ . - (1 | donor_ref) + (1 + z_s1_share | donor_ref)),
        data = cc, family = binomial, control = CTRL),
  silent = TRUE
)

if (!inherits(m5, "try-error")) {
  anova(m5, m4) |> broom::tidy() |>
    transmute(Model = c("M4 - random intercept", "M5 - + random slope"),
              npar, AIC = round(AIC), BIC = round(BIC),
              `LRT chi-sq` = round(statistic, 2), df,
              p = round(p.value, 4)) |>
    show_tbl("Does the share effect differ across donors?")
} else {
  tibble(Result = "The random-slope model did not converge; the effect is treated as constant across donors.") |>
    show_tbl()
}
```

# H2 — Does a larger share predict interruption?

**The prediction.** Substitutability says a donor supplying a large share cannot
easily be replaced, so its threat is credible and it uses it. The coefficient
should be **positive**.

```{r}
#| label: impute
#| cache: true
set.seed(20260824)

imp <- model_df |>
  select(all_of(MODEL_VARS)) |>
  mice(m = 5, method = "pmm", printFlag = FALSE, seed = 20260824)

fit_pooled <- function(formula, imputations = imp) {
  map(seq_len(imputations$m), \(i)
      glmer(formula, data = complete(imputations, i),
            family = binomial, control = CTRL)) |>
    pool()
}
```

```{r}
#| label: h2
#| cache: true
pool_h2 <- fit_pooled(f4)
tidy_pooled(pool_h2) |> show_tbl("H2: full model, pooled across imputations")
```

# H3 — Fragmentation, and which mechanism it supports

The two mechanisms disagree here and only here:

| Mechanism | Prediction on the number of donors |
|---|---|
| Substitutability | **Negative** — more donors, easier to replace, less interruption |
| Signalling | **Positive** — more donors, larger audience, more interruption |

`s1_share` and `s2_hhi_net` correlate at about −0.87 by construction, so they
never enter together. H3 is estimated on pool size, with concentration reported
as a second measure of the same idea.

```{r}
#| label: h3
#| cache: true
f_h3  <- update(f4, . ~ . - z_s1_share + z_s3_n_donors)
f_h3b <- update(f4, . ~ . - z_s1_share + z_s2_hhi_net)

pool_h3  <- fit_pooled(f_h3)
pool_h3b <- fit_pooled(f_h3b)

bind_rows(
  tidy_pooled(pool_h3,  "z_s3_n_donors") |> mutate(Measure = "Number of donors"),
  tidy_pooled(pool_h3b, "z_s2_hhi_net")  |> mutate(Measure = "Concentration of the others")
) |>
  relocate(Measure) |>
  show_tbl("H3: two measures of the same idea")
```

# The focal estimates together

```{r}
#| label: fig-coefs
#| fig-cap: "The focal coefficient from each model, as odds ratios. Above 1 means the variable raises the odds of interruption; below 1, lowers them."
#| fig-height: 3.2
grab <- function(pooled, term_name, label) {
  s <- summary(pooled, conf.int = TRUE) |> as_tibble()
  names(s)[names(s) == "2.5 %"] <- "lo"; names(s)[names(s) == "97.5 %"] <- "hi"
  s |> filter(term == term_name) |> mutate(Model = label)
}

bind_rows(
  grab(pool_h2,  "z_s1_share",    "H2 - donor share"),
  grab(pool_h3,  "z_s3_n_donors", "H3 - number of donors"),
  grab(pool_h3b, "z_s2_hhi_net",  "H3 - others' concentration"),
  grab(pool_h2,  "donor_typeMultilateral", "Benchmark - multilateral")
) |>
  mutate(Model = fct_rev(fct_inorder(Model))) |>
  ggplot(aes(exp(estimate), Model)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = exp(lo), xmax = exp(hi)), size = 0.5) +
  scale_x_log10() +
  labs(x = "Odds ratio (log scale)", y = NULL)
```

# What the estimates mean, in probabilities

Odds ratios are hard to read. This converts the H2 estimate into predicted
probabilities of interruption across the observed range of donor share, holding
everything else at its mean and the random effects at zero.

```{r}
#| label: fig-predicted
#| fig-cap: "Predicted probability of interruption across the range of donor share, for an average country-year. The ribbon is the 95% interval from the fixed effects."
#| fig-height: 3.8
m_pred <- glmer(f4, data = cc, family = binomial, control = CTRL)

newdat <- tibble(
  z_s1_share = seq(min(cc$z_s1_share), max(cc$z_s1_share), length.out = 60)
) |>
  crossing(donor_type = factor(c("Bilateral", "Multilateral"),
                               levels = levels(cc$donor_type))) |>
  mutate(z_log_commitment = 0, coup = 0, election_year = 0, year_c = 5,
         z_voice_account = 0, z_corruption_ctrl = 0, z_log_aid_dep = 0,
         z_log_gdp_pc = 0, z_net_degree = 0)

X  <- model.matrix(delete.response(terms(m_pred)), newdat)
b  <- fixef(m_pred)
se <- sqrt(diag(X %*% vcov(m_pred) %*% t(X)))

newdat |>
  mutate(fit = plogis(X %*% b)[, 1],
         lo  = plogis(X %*% b - 1.96 * se)[, 1],
         hi  = plogis(X %*% b + 1.96 * se)[, 1],
         # back to the original scale for a readable axis
         share = z_s1_share * sd(d$s1_share, na.rm = TRUE) +
                 mean(d$s1_share, na.rm = TRUE)) |>
  filter(between(share, 0, 1)) |>
  ggplot(aes(share, fit, colour = donor_type, fill = donor_type)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = PAL) +
  scale_fill_manual(values = PAL) +
  scale_x_continuous(labels = label_percent()) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "The donor's share of all money promised to that country",
       y = "Predicted probability of interruption", colour = NULL, fill = NULL)
```

# Which donors depart from the model

The random intercepts say how far each donor sits above or below what its
observable characteristics predict. A donor high on this scale interrupts more
than the model expects of a donor like it.

```{r}
#| label: fig-ranef
#| fig-cap: "Donor random intercepts with 95% intervals. Zero is the average donor after controls; positive means more interruption than predicted."
#| fig-height: 4.5
re <- ranef(m_pred, condVar = TRUE)$donor_ref
pv <- attr(re, "postVar")[1, 1, ]

lookup <- cc |> distinct(donor_ref, donor, donor_type)

tibble(donor_ref = rownames(re), est = re[, 1], se = sqrt(pv)) |>
  left_join(lookup, by = "donor_ref") |>
  mutate(donor = fct_reorder(str_trunc(donor, 32), est)) |>
  ggplot(aes(est, donor, colour = donor_type)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = est - 1.96 * se, xmax = est + 1.96 * se), size = 0.42) +
  scale_colour_manual(values = PAL) +
  labs(x = "Donor random intercept (log-odds)", y = NULL, colour = NULL)
```

# Save

```{r}
#| label: save
saveRDS(list(m1 = m1, m4 = m4, pool_h2 = pool_h2, pool_h3 = pool_h3,
             pool_h3b = pool_h3b, cc = cc),
        "../data/processed/models_h2_h3.rds")
```
