---
title: "H1 — Does structural position predict better than institutional type?"
subtitle: "Repeated grouped cross-validation across predictor sets and model families"
author: "Gaspar Vigneaux"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    theme: cosmo
    fontsize: 1.02em
    linestretch: 1.5
    max-width: 1100px
    fig-width: 8
    fig-height: 4.5
    fig-align: center
    embed-resources: true
execute:
  warning: false
  message: false
---

**H1.** *A donor's structural position predicts interruption better than whether
the donor is bilateral or multilateral.*

05_models answered what the mechanism is and in which direction it runs.
This document answers a different question — **which set of variables generalises
to operations the model has never seen.**

**This version replaces a single train-test split.** An earlier draft compared
the two predictor sets on one split and reported a difference of 0.009 in AUC.
With 436 test rows the standard error on an AUC near 0.71 is about 0.027, so
that difference was a fifth of one standard error and the comparison could not
distinguish anything. The comparison is now made across
**25 resamples**, and what is reported is the **paired difference** rather than
two separate point estimates.

```{r}
#| label: setup
library(tidyverse)
library(glmnet)
library(rpart)
library(rpart.plot)
library(ranger)
library(gbm)
library(pROC)
library(kableExtra)
library(scales)
library(patchwork)
library(conflicted)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
conflict_prefer("slice", "dplyr")

d <- read_csv("../data/processed/model_data_with_network.csv", show_col_types = FALSE)

theme_set(
  theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold", size = 11),
          plot.subtitle = element_text(colour = "grey35", size = 9.5),
          legend.position = "bottom")
)

PAL_SET <- c("Controls only" = "grey60",
             "Institutional type" = "#C0504D",
             "Structural position" = "#4F6D8C",
             "Structural (components)" = "#8064A2",
             "Both" = "#7B9E6F",
             "Donor identity (ceiling)" = "grey25")

show_tbl <- function(df, caption = NULL) {
  df |>
    knitr::kable(caption = caption, digits = 3, align = "l") |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE, position = "left", font_size = 13)
}
```

# The rules of the comparison

**One thing changes at a time.** Every model sees the same observations, the
same controls and the same estimation procedure. Only the focal variables
differ. Otherwise a difference in performance would measure the specification
rather than the hypothesis.

**Folds are grouped by activity.** An activity contributes several years.
Splitting at random would place 2018 and 2019 of the same operation on opposite
sides, so a model would meet an operation's own history in training and be
scored on it in testing.

**Complete cases, not imputation.** Multiple imputation inside a resampling loop
requires pooling predictions across imputed datasets, which is rarely done
transparently. The cost in sample size is reported below.

**Hyperparameters are tuned once, then held fixed across every fold and set.**
Tuning inside each fold would be stricter, but it is also where the runtime goes
and — more importantly — the point here is to compare *predictor sets*, and
fixing the settings guarantees that no set is favoured by having been tuned
harder. The mild optimism this introduces applies equally to all of them, so it
cannot generate a difference between them.

**That tuning does breach one rule, and the breach is bounded.** The course
principle is that the held-out data are used once, and here the cross-validated
penalty and boosting length are chosen on the full complete-case sample, which
later supplies every test fold. The consequence is that the **absolute** areas
reported below are somewhat optimistic and should not be read as an estimate of
how these models would perform on genuinely new data. The consequence for the
**comparison** is nil, because the same optimism is applied to every predictor
set by construction, and H1 is a question about the difference rather than about
the level. The same argument covers the principal components, whose penalty and
boosting length are set from a single rotation on the full sample even though the
rotation used for prediction is refitted inside each fold.

**On the software.** The estimation packages here are `glmnet` for the penalised
model, `rpart` for the tree, `ranger` for the forest, `gbm` for the boosted
ensemble and `pROC` for the ROC machinery. The methods are those of the course;
these are implementations of them, chosen for speed on repeated resampling.
`caret`, the framework the course uses to wrap cross-validation, is deliberately
not used: its resampling helpers partition rows, and the partition here has to
respect activities, so the loop is written out instead. Writing it out also makes
the grouping visible rather than hiding it behind a control object.

```{r}
#| label: prepare
set.seed(20260824)

model_df <- d |>
  mutate(
    log_aid_dep = if_else(aid_dependence > 0, log(pmax(aid_dependence, 1e-6)), NA_real_),
    donor_type  = factor(x1_donor_type, levels = c("Bilateral", "Multilateral")),
    donor_f     = factor(donor_ref),
    year_c      = year - 2014,
    y           = factor(interrupted_flag, levels = c(0, 1), labels = c("no", "yes"))
  )

CONTROLS <- c("voice_account", "corruption_ctrl", "log_aid_dep", "log_gdp_pc",
              "log_commitment", "coup", "election_year", "year_c")
STRUCT   <- c("s1_share", "s2_hhi_net", "s3_n_donors")
INSTIT   <- "donor_type"

analysis_df <- model_df |>
  select(iatiidentifier, y, all_of(c(CONTROLS, STRUCT, INSTIT)), donor_f) |>
  drop_na() |>
  # donor_f is built before the complete-case filter. Without re-levelling, the
  # design matrix carries an all-zero column for each donor the filter removes,
  # which inflates the parameter count reported for the ceiling model.
  mutate(donor_f = droplevels(donor_f))

tibble(
  Stage = c("All activity-years", "Complete cases"),
  Rows = c(nrow(model_df), nrow(analysis_df)),
  Activities = c(n_distinct(model_df$iatiidentifier), n_distinct(analysis_df$iatiidentifier))
) |>
  mutate(`% retained` = round(100 * Rows / Rows[1])) |>
  show_tbl("What the complete-case restriction costs")
```

# The six predictor sets

```{r}
#| label: sets
sets <- list(
  "Controls only"            = CONTROLS,
  "Institutional type"       = c(CONTROLS, INSTIT),
  "Structural position"      = c(CONTROLS, STRUCT),
  "Both"                     = c(CONTROLS, INSTIT, STRUCT),
  "Donor identity (ceiling)" = c(CONTROLS, "donor_f")
)

tibble(
  `Predictor set` = names(sets),
  Adds = c("nothing - the floor every other model must beat",
           "bilateral or multilateral",
           "donor share, concentration of others, number of donors",
           "both of the above",
           "which donor it is, as a factor")
) |>
  show_tbl("What each model sees")
```

- **"Controls only" is the floor.** If a focal set does not beat it, that set
  carries nothing the recipient-side controls do not already have.
- **"Donor identity" is the ceiling**. The null model in
  05_models put 30 per cent of the outcome variance between donors, so a
  model that simply knows *which donor* should predict well. It is not a
  competitor to the hypotheses — it is the scale on which they are read. If
  structural position recovers much of the gap between floor and ceiling, it is
  capturing what makes donors differ; if it does not, whatever separates donors
  lies elsewhere.

A sixth set using principal components of the structural variables is added
after the rotation is fitted, below.

# Resampling design

```{r}
#| label: folds
N_REPEATS <- 5
N_FOLDS   <- 5

acts <- unique(analysis_df$iatiidentifier)

fold_map <- map_dfr(seq_len(N_REPEATS), \(r) {
  set.seed(20260824 + r)
  tibble(repeat_id = r,
         iatiidentifier = acts,
         fold = sample(rep_len(seq_len(N_FOLDS), length(acts))))
})

resamples <- fold_map |> distinct(repeat_id, fold) |> mutate(id = row_number())

tibble(
  Repeats = N_REPEATS, Folds = N_FOLDS,
  Resamples = nrow(resamples),
  `Activities per test fold` = round(length(acts) / N_FOLDS),
  `Rows per test fold` = round(nrow(analysis_df) / N_FOLDS)
) |>
  show_tbl("The resampling scheme")
```

- **The same folds are used by every model.** That is what makes the comparison
  paired: two predictor sets are scored on identical held-out observations, so
  the fold-to-fold noise they share cancels out of their difference. A paired
  difference is far better determined than either level, which is precisely the
  problem with the single split this replaces.

# Tuning, once

```{r}
#| label: tune
#| cache: true
mx <- function(df, vars) model.matrix(~ . - 1, data = df |> select(all_of(vars)))

# One cross-validated lambda per predictor set, on the full complete-case data.
lambdas <- map(sets, \(v) {
  cv <- cv.glmnet(mx(analysis_df, v), analysis_df$y, family = "binomial",
                  alpha = 0, nfolds = 10)
  cv$lambda.min
})

# One boosting length per predictor set, chosen by gbm's own cross-validation.
n_trees <- map(sets, \(v) {
  m <- gbm(as.numeric(analysis_df$y == "yes") ~ .,
           data = analysis_df |> select(all_of(v)),
           distribution = "bernoulli", n.trees = 1500, interaction.depth = 3,
           shrinkage = 0.01, n.minobsinnode = 20, cv.folds = 5, verbose = FALSE)
  gbm.perf(m, method = "cv", plot.it = FALSE)
})

tibble(`Predictor set` = names(sets),
       `Ridge lambda` = round(unlist(lambdas), 4),
       `Boosting trees` = unlist(n_trees)) |>
  show_tbl("Settings fixed before resampling begins")
```

# Fitting

```{r}
#| label: fit-functions
auc_of <- function(truth, prob) {
  as.numeric(pROC::auc(pROC::roc(truth, prob, quiet = TRUE,
                                 levels = c("no", "yes"), direction = "<")))
}
brier_of <- function(truth, prob) mean((as.numeric(truth == "yes") - prob)^2)

# Sensitivity, specificity and their mean at one cut-point. Computed here rather
# than in a separate document because the predicted probabilities exist only
# inside this loop: a threshold metric cannot be recovered from a stored area.
thr_of <- function(truth, prob, thr) {
  hit <- prob >= thr
  pos <- truth == "yes"
  sens <- sum(hit & pos) / sum(pos)
  spec <- sum(!hit & !pos) / sum(!pos)
  c(sens = sens, spec = spec, bal = (sens + spec) / 2, flagged = mean(hit))
}

fit_one <- function(tr, te, vars, lambda, ntree) {
  f <- as.formula(paste("y ~", paste(vars, collapse = " + ")))

  # Ridge rather than lasso: the structural variables are correlated by
  # construction, and lasso is unstable with correlated groups.
  g  <- glmnet(mx(tr, vars), tr$y, family = "binomial", alpha = 0, lambda = lambda)
  p_glm <- as.numeric(predict(g, newx = mx(te, vars), type = "response"))

  t0 <- rpart(f, data = tr |> select(y, all_of(vars)), method = "class",
              control = rpart.control(cp = 0.001, xval = 10))
  cpt  <- t0$cptable
  best <- cpt[which.min(cpt[, "xerror"]), ]
  cp1  <- cpt[which(cpt[, "xerror"] <= best[["xerror"]] + best[["xstd"]])[1], "CP"]
  p_tree <- predict(prune(t0, cp = cp1), newdata = te, type = "prob")[, "yes"]

  rf <- ranger(f, data = tr |> select(y, all_of(vars)), probability = TRUE,
               num.trees = 800, seed = 20260824)
  p_rf <- predict(rf, te)$predictions[, "yes"]

  gb <- gbm(as.numeric(tr$y == "yes") ~ ., data = tr |> select(all_of(vars)),
            distribution = "bernoulli", n.trees = ntree, interaction.depth = 3,
            shrinkage = 0.01, n.minobsinnode = 20, verbose = FALSE)
  p_gb <- predict(gb, newdata = te |> select(all_of(vars)), n.trees = ntree,
                  type = "response")

  ps <- list(p_glm, p_tree, p_rf, p_gb)

  # The prevalence cut-point comes from the training fold. Taking it from the
  # test fold would let the held-out data set its own decision rule.
  prev  <- mean(tr$y == "yes")
  m_50  <- map(ps, \(p) thr_of(te$y, p, 0.5))
  m_prv <- map(ps, \(p) thr_of(te$y, p, prev))

  tibble(Family = c("Penalised logistic", "Classification tree",
                    "Random forest", "Gradient boosting"),
         AUC   = map_dbl(ps, \(p) auc_of(te$y, p)),
         Brier = map_dbl(ps, \(p) brier_of(te$y, p)),
         `Sensitivity (0.5)`        = map_dbl(m_50,  "sens"),
         `Specificity (0.5)`        = map_dbl(m_50,  "spec"),
         `Balanced accuracy (0.5)`  = map_dbl(m_50,  "bal"),
         `Share flagged (0.5)`      = map_dbl(m_50,  "flagged"),
         `Sensitivity (prev)`       = map_dbl(m_prv, "sens"),
         `Specificity (prev)`       = map_dbl(m_prv, "spec"),
         `Balanced accuracy (prev)` = map_dbl(m_prv, "bal"))
}
```

```{r}
#| label: pca
# The rotation is learned inside each fold's training data, never on the whole
# sample: fitting it once on everything would let the test folds influence their
# own representation.
add_pca <- function(tr, te) {
  rot <- prcomp(tr |> select(all_of(STRUCT)), center = TRUE, scale. = TRUE)
  list(tr = bind_cols(tr, as_tibble(predict(rot, tr |> select(all_of(STRUCT))))),
       te = bind_cols(te, as_tibble(predict(rot, te |> select(all_of(STRUCT))))),
       rot = rot)
}

PCS <- c("PC1", "PC2")
sets[["Structural (components)"]] <- c(CONTROLS, PCS)

# The components need their own tuning constants; taken from a single rotation
# on the full data purely to set lambda and the boosting length.
tmp <- add_pca(analysis_df, analysis_df)
lambdas[["Structural (components)"]] <-
  cv.glmnet(mx(tmp$tr, c(CONTROLS, PCS)), analysis_df$y, family = "binomial",
            alpha = 0, nfolds = 10)$lambda.min
n_trees[["Structural (components)"]] <- 600
```

```{r}
#| label: run
#| cache: true
results <- map_dfr(seq_len(nrow(resamples)), \(i) {
  r  <- resamples$repeat_id[i]; k <- resamples$fold[i]
  ids <- fold_map |> filter(repeat_id == r)
  tr_ids <- ids$iatiidentifier[ids$fold != k]

  tr <- analysis_df |> filter(iatiidentifier %in% tr_ids)
  te <- analysis_df |> filter(!iatiidentifier %in% tr_ids)
  pc <- add_pca(tr, te)

  map_dfr(names(sets), \(s) {
    v <- sets[[s]]
    use_pc <- any(v %in% PCS)
    fit_one(if (use_pc) pc$tr else tr,
            if (use_pc) pc$te else te,
            v, lambdas[[s]], n_trees[[s]]) |>
      mutate(`Predictor set` = s, .before = 1)
  }) |>
    mutate(resample = i, repeat_id = r, fold = k)
})

nrow(results)
```

# Results

```{r}
#| label: summary
# sd() must be computed before AUC is overwritten: inside summarise(), a later
# expression sees the new column, not the original one, and returns NA.
summ <- results |>
  summarise(`AUC SD` = sd(AUC), Brier = mean(Brier), AUC = mean(AUC),
            .by = c(`Predictor set`, Family)) |>
  relocate(AUC, .after = Family) |>
  arrange(desc(AUC))

summ |> show_tbl("Mean performance across 25 resamples")
```

## What each set costs to buy that

Accuracy on its own does not settle a comparison between specifications. A set
that buys a gain with many parameters is not doing the same thing as one that
buys a smaller gain with a single distinction, and the course material is
explicit that complexity is tracked alongside performance rather than after it.

```{r}
#| label: complexity
# Counted from the design matrix glmnet actually receives, not from the number
# of variable names: a factor contributes one column per level it retains.
n_params <- map_int(sets, \(v) {
  if (any(v %in% PCS)) length(CONTROLS) + length(PCS) else ncol(mx(analysis_df, v))
})

base_p <- n_params[["Controls only"]]

summ |>
  filter(Family != "Classification tree") |>
  summarise(AUC = mean(AUC), .by = `Predictor set`) |>
  mutate(Parameters = n_params[`Predictor set`],
         `Gain over controls` = AUC - AUC[`Predictor set` == "Controls only"],
         `Parameters added` = Parameters - base_p,
         `Gain per parameter` = if_else(`Parameters added` > 0,
                                        `Gain over controls` / `Parameters added`,
                                        NA_real_)) |>
  arrange(desc(AUC)) |>
  show_tbl("Performance against complexity, averaged over the three families that discriminate")
```

- **Read the last column, not the third.** The ceiling reaches the highest area,
  but it does so by spending a parameter on every donor. Institutional type
  spends one binary distinction. If the second buys a comparable share of the
  gain at a fraction of the cost, then what separates donors is closer to a
  category than to an identity — and that bears directly on H1, which proposes
  a third possibility, position, that must earn its own parameters.

```{r}
#| label: fig-auc
#| fig-cap: "AUC across resamples. Each point is one fold; the bar is the mean. Spread within a predictor set shows how much a single split would have moved the answer."
#| fig-height: 4.6
results |>
  mutate(`Predictor set` = fct_reorder(`Predictor set`, AUC)) |>
  ggplot(aes(AUC, `Predictor set`, colour = `Predictor set`)) +
  geom_vline(xintercept = 0.5, linetype = "dashed", colour = "grey60") +
  geom_jitter(height = 0.16, alpha = 0.3, size = 1) +
  stat_summary(fun = mean, geom = "point", size = 3.2) +
  scale_colour_manual(values = PAL_SET, guide = "none") +
  facet_wrap(~Family) +
  labs(y = NULL, title = "Where a single split would have landed")
```

# The paired comparison, which is the actual test

Two predictor sets scored on the **same** fold share that fold's difficulty. The
difference between them removes it. This is why the paired difference is
informative where the two levels separately are not.

```{r}
#| label: paired
paired <- results |>
  select(`Predictor set`, Family, resample, AUC) |>
  pivot_wider(names_from = `Predictor set`, values_from = AUC) |>
  mutate(
    `Structural - Institutional` = `Structural position` - `Institutional type`,
    `Structural - Controls`      = `Structural position` - `Controls only`,
    `Institutional - Controls`   = `Institutional type`  - `Controls only`,
    `Ceiling - Controls`         = `Donor identity (ceiling)` - `Controls only`
  )

# The 25 resamples are not independent: five repeats reuse the same rows, so the
# spread across folds understates how uncertain the mean difference is. The
# Nadeau-Bengio correction adds the test-to-train size ratio to the variance,
# which at five folds inflates the standard error by a factor of about 2.7.
RHO <- (1 / N_FOLDS) / (1 - 1 / N_FOLDS)

paired |>
  select(Family, resample, starts_with("Structural -"), starts_with("Institutional -"),
         starts_with("Ceiling -")) |>
  pivot_longer(-c(Family, resample), names_to = "Comparison", values_to = "diff") |>
  summarise(
    `Mean difference` = mean(diff),
    # A percentile range of the observed folds. This describes dispersion, not
    # the precision of the mean, and was previously mislabelled as an interval.
    `95% of folds` = paste0("[", round(quantile(diff, 0.025), 3), ", ",
                            round(quantile(diff, 0.975), 3), "]"),
    # An interval on the mean difference, which is what the comparison asks about.
    `95% CI, corrected` = paste0(
      "[", round(mean(diff) - 1.96 * sd(diff) * sqrt(1 / n() + RHO), 3), ", ",
      round(mean(diff) + 1.96 * sd(diff) * sqrt(1 / n() + RHO), 3), "]"),
    `Wins` = paste0(round(100 * mean(diff > 0)), "%"),
    .by = c(Comparison, Family)
  ) |>
  arrange(Comparison, desc(`Mean difference`)) |>
  show_tbl("Paired differences in AUC across the 25 resamples")
```

```{r}
#| label: fig-paired
#| fig-cap: "Distribution of the paired AUC difference across resamples. An interval crossing zero means the two sets cannot be separated."
#| fig-height: 4
paired |>
  select(Family, resample, `Structural - Institutional`, `Structural - Controls`,
         `Institutional - Controls`, `Ceiling - Controls`) |>
  pivot_longer(-c(Family, resample), names_to = "Comparison", values_to = "diff") |>
  mutate(Comparison = fct_rev(fct_inorder(Comparison))) |>
  ggplot(aes(diff, Comparison)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_jitter(width = 0, height = 0.15, alpha = 0.35, size = 1.1, colour = "grey40") +
  stat_summary(fun = mean, geom = "point", size = 3, colour = "firebrick") +
  facet_wrap(~Family) +
  labs(x = "Difference in AUC", y = NULL)
```

```{r}
#| label: h1-verdict
#| echo: false
d_si <- paired$`Structural - Institutional`
d_sc <- paired$`Structural - Controls`
d_ceil <- paired$`Ceiling - Controls`
```

**The answer to H1.** Across all families and resamples, structural position
differs from institutional type by a mean of `r round(mean(d_si), 3)` AUC, with
95 per cent of folds falling between `r round(quantile(d_si, 0.025), 3)` and
`r round(quantile(d_si, 0.975), 3)`, and a win rate of
`r round(100 * mean(d_si > 0))` per cent of folds.

**The two interval columns above answer different questions.** The percentile
range describes how far the difference moves from fold to fold. The corrected
interval is the one that bears on the hypothesis: it asks how precisely the mean
difference is known, and it is wider because five repeats over the same rows do
not supply twenty-five independent measurements. On that interval the two focal
sets cannot be separated in either tree ensemble, while the penalised logistic
favours institutional type. Neither reading supports H1.

Read that against the falsification condition in H1, and against two reference
points: structural position beats the controls-only floor by
`r round(mean(d_sc), 3)`, and knowing the donor's identity beats it by
`r round(mean(d_ceil), 3)`. **The second number is the scale.** If the
hypothesis's variable recovers only a small fraction of what donor identity
recovers, then whatever makes donors differ is mostly not their structural
position.

# A second check on the single split

DeLong's test compares two correlated AUCs computed on the same held-out set.
It does not replace resampling — one split remains one split — but it reports a
p-value on the comparison the earlier draft made, using the same machinery a
reader familiar with ROC analysis would expect.

```{r}
#| label: delong
#| cache: true
set.seed(20260824)
hold_acts <- sample(acts, floor(0.25 * length(acts)))
tr1 <- analysis_df |> filter(!iatiidentifier %in% hold_acts)
te1 <- analysis_df |> filter(iatiidentifier %in% hold_acts)

rf_s <- ranger(as.formula(paste("y ~", paste(sets[["Structural position"]], collapse = "+"))),
               data = tr1, probability = TRUE, num.trees = 800, seed = 1)
rf_i <- ranger(as.formula(paste("y ~", paste(sets[["Institutional type"]], collapse = "+"))),
               data = tr1, probability = TRUE, num.trees = 800, seed = 1)

roc_s <- roc(te1$y, predict(rf_s, te1)$predictions[, "yes"], quiet = TRUE,
             levels = c("no", "yes"), direction = "<")
roc_i <- roc(te1$y, predict(rf_i, te1)$predictions[, "yes"], quiet = TRUE,
             levels = c("no", "yes"), direction = "<")

dl <- roc.test(roc_s, roc_i, method = "delong")

tibble(`AUC structural` = round(as.numeric(roc_s$auc), 3),
       `AUC institutional` = round(as.numeric(roc_i$auc), 3),
       Z = round(dl$statistic, 3),
       p = round(dl$p.value, 4)) |>
  show_tbl("DeLong test on a single held-out set")
```

# Does the choice of metric decide the answer?

Area under the curve and the Brier score are both threshold-free. That is a
choice, and it could be the thing producing the verdict rather than the data.
Two ROC curves can enclose the same area and still **cross**, one predictor set
ranking better where specificity is high and the other where sensitivity is
high. The area would average that difference away.

The fitting loop above now also scores every fold at two cut-points: the
conventional 0.5, and the training prevalence, which is the natural choice when
the outcome is not balanced. No model is refitted — these are the same
predictions, read a second way.

```{r}
#| label: threshold-summary
results |>
  summarise(across(c(`Sensitivity (0.5)`, `Specificity (0.5)`,
                     `Balanced accuracy (0.5)`, `Share flagged (0.5)`,
                     `Sensitivity (prev)`, `Specificity (prev)`,
                     `Balanced accuracy (prev)`), mean),
            .by = c(`Predictor set`, Family)) |>
  arrange(desc(`Balanced accuracy (prev)`)) |>
  show_tbl("Threshold metrics, averaged over the 25 resamples")
```

- **At 0.5 the metric measures nothing about the predictor sets.** Interruption
  runs at about a fifth of activity-years, and no model puts a fifth of its test
  rows above a half. Read the share-flagged column: every set flags far less
  than the true rate, so balanced accuracy sits just above 0.5 for all of them
  and the sets cannot be told apart. This is the accuracy paradox, and it is why
  the comparison is not conducted on a threshold metric — now shown rather than
  asserted.
- **At the prevalence cut-point the metric does separate**, and the ordering it
  produces is the one the area already gave.

```{r}
#| label: threshold-paired
results |>
  select(`Predictor set`, Family, resample,
         `Balanced accuracy (0.5)`, `Balanced accuracy (prev)`) |>
  pivot_longer(starts_with("Balanced"), names_to = "Threshold", values_to = "ba") |>
  pivot_wider(names_from = `Predictor set`, values_from = ba) |>
  mutate(diff = `Structural position` - `Institutional type`) |>
  summarise(`Mean difference` = mean(diff),
            `95% CI` = paste0("[", round(quantile(diff, 0.025), 3), ", ",
                              round(quantile(diff, 0.975), 3), "]"),
            Wins = paste0(round(100 * mean(diff > 0)), "%"),
            .by = Threshold) |>
  show_tbl("Structural position minus institutional type, in balanced accuracy")
```

- Compare against the area, which returned a mean paired difference of
  `r round(mean(d_si), 3)` with a win rate of `r round(100 * mean(d_si > 0))`
  per cent. **The verdict does not depend on the metric.**

## Do the two curves cross?

If they do, the two sets are qualitatively different at the same area and
reporting only the area conceals it. The difference in sensitivity is read at
fixed levels of specificity, where a crossing appears as a change of sign down
the column.

**This cannot be read off the DeLong split.** That split holds around ninety
interruptions, so a handful of cases moves sensitivity by several points and any
sign change it produces is unreadable. Out-of-fold predictions are pooled across
the 25 resamples instead, which is the same evidence the rest of the comparison
rests on.

```{r}
#| label: crossing-preds
#| cache: true
focal <- c("Structural position", "Institutional type")

oof <- map_dfr(seq_len(nrow(resamples)), \(i) {
  r <- resamples$repeat_id[i]; k <- resamples$fold[i]
  ids <- fold_map |> filter(repeat_id == r)
  tr_ids <- ids$iatiidentifier[ids$fold != k]

  tr <- analysis_df |> filter(iatiidentifier %in% tr_ids)
  te <- analysis_df |> filter(!iatiidentifier %in% tr_ids)

  map_dfr(focal, \(s) {
    v <- sets[[s]]
    f <- as.formula(paste("y ~", paste(v, collapse = " + ")))

    g  <- glmnet(mx(tr, v), tr$y, family = "binomial", alpha = 0,
                 lambda = lambdas[[s]])
    rf <- ranger(f, data = tr |> select(y, all_of(v)), probability = TRUE,
                 num.trees = 800, seed = 20260824)
    gb <- gbm(as.numeric(tr$y == "yes") ~ ., data = tr |> select(all_of(v)),
              distribution = "bernoulli", n.trees = n_trees[[s]],
              interaction.depth = 3, shrinkage = 0.01, n.minobsinnode = 20,
              verbose = FALSE)

    tibble(`Predictor set` = s,
           Family = rep(c("Penalised logistic", "Random forest",
                          "Gradient boosting"), each = nrow(te)),
           truth = rep(te$y, 3),
           p = c(as.numeric(predict(g, newx = mx(te, v), type = "response")),
                 predict(rf, te)$predictions[, "yes"],
                 predict(gb, newdata = te |> select(all_of(v)),
                         n.trees = n_trees[[s]], type = "response")))
  })
})
```

```{r}
#| label: crossing
sens_at <- function(set_name, fam, sp) {
  x <- oof |> filter(`Predictor set` == set_name, Family == fam)
  r <- roc(x$truth, x$p, quiet = TRUE, levels = c("no", "yes"), direction = "<")
  coords(r, x = sp, input = "specificity", ret = "sensitivity")[1, 1]
}

expand_grid(Family = c("Gradient boosting", "Random forest", "Penalised logistic"),
            Specificity = c(0.95, 0.90, 0.80, 0.70, 0.50)) |>
  mutate(`Sensitivity, structural` =
           map2_dbl(Family, Specificity, \(f, s) sens_at("Structural position", f, s)),
         `Sensitivity, institutional` =
           map2_dbl(Family, Specificity, \(f, s) sens_at("Institutional type", f, s)),
         Difference = `Sensitivity, structural` - `Sensitivity, institutional`) |>
  show_tbl("Sensitivity of each focal set at matched specificity, pooled across resamples")
```

- A sign that flips down the Difference column would be a crossing. Judge any
  such flip against the fold-to-fold standard deviation of the area, which is
  about four points: a difference smaller than that is noise wearing the shape
  of a pattern.
- Pooling mixes curves fitted on different training sets, which is why this is
  read as a diagnostic on the shape of the trade-off rather than as an estimate
  of any one model's operating characteristics.

# Which family, and does flexibility help?

```{r}
#| label: families
results |>
  summarise(`SD` = sd(AUC), `Mean Brier` = mean(Brier), `Mean AUC` = mean(AUC),
            .by = Family) |>
  relocate(`Mean AUC`, .after = Family) |>
  arrange(desc(`Mean AUC`)) |>
  show_tbl("Performance by model family, averaged over sets and resamples")
```

- If the flexible families beat the penalised logistic substantially, the
  relationship carries interactions or non-linearities a linear model cannot
  represent — a substantive claim about how leverage operates.
- If they do not, the interpretable model is justified **on evidence**, which
  the course material identifies as the social-science argument.

**Not fitted: neural networks.** They occupy the same flexibility tier as the
forest and the boosted trees and are excluded on two grounds. Around 1,400
training rows against a dozen predictors is far too little for a network to
learn anything the tree ensembles do not already capture; and an uninterpretable
model that does not predict better has nothing to offer a thesis whose claims
are substantive rather than operational.

# Variable importance, averaged across resamples

Importance from a single fit on a single split is unstable at this sample
size. This averages permutation importance over the same 25 resamples.

```{r}
#| label: importance
#| cache: true
imp_lbl <- c(s1_share = "Donor share", s2_hhi_net = "Concentration of others",
             s3_n_donors = "Donors in country", donor_type = "Institutional type",
             log_aid_dep = "Aid dependence", log_commitment = "Operation size",
             voice_account = "Voice and accountability",
             corruption_ctrl = "Control of corruption",
             log_gdp_pc = "GDP per capita", coup = "Coup",
             election_year = "Election", year_c = "Year")

vars_both <- sets[["Both"]]

imp <- map_dfr(seq_len(nrow(resamples)), \(i) {
  r <- resamples$repeat_id[i]; k <- resamples$fold[i]
  ids <- fold_map |> filter(repeat_id == r)
  tr <- analysis_df |> filter(iatiidentifier %in% ids$iatiidentifier[ids$fold != k])
  m <- ranger(as.formula(paste("y ~", paste(vars_both, collapse = "+"))),
              data = tr, probability = TRUE, num.trees = 800,
              importance = "permutation", seed = 20260824)
  tibble(Variable = names(m$variable.importance),
         Importance = m$variable.importance, resample = i)
})

imp_summary <- imp |>
  summarise(mean_imp = mean(Importance), sd_imp = sd(Importance), .by = Variable) |>
  mutate(Label = coalesce(imp_lbl[Variable], Variable),
         Focal = case_when(Variable %in% STRUCT ~ "Structural position",
                           Variable == INSTIT   ~ "Institutional type",
                           TRUE ~ "Control")) |>
  arrange(desc(mean_imp))

imp_summary |>
  transmute(Variable = Label, Role = Focal,
            `Mean importance` = round(mean_imp, 5),
            `SD` = round(sd_imp, 5)) |>
  show_tbl("Permutation importance, averaged over 25 resamples")
```

```{r}
#| label: fig-importance
#| fig-cap: "Permutation importance averaged across resamples, with one standard deviation. Higher means the model loses more accuracy when that variable is scrambled."
#| fig-height: 4.2
imp_summary |>
  mutate(Label = fct_reorder(Label, mean_imp)) |>
  ggplot(aes(mean_imp, Label, colour = Focal)) +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_pointrange(aes(xmin = mean_imp - sd_imp, xmax = mean_imp + sd_imp), size = 0.4) +
  scale_colour_manual(values = c("Structural position" = "#4F6D8C",
                                 "Institutional type" = "#C0504D",
                                 "Control" = "grey55")) +
  labs(x = "Mean permutation importance", y = NULL, colour = NULL)
```

- Compare this ranking against the coefficients in 05_models. **Agreement
  between a method that assumes a functional form and one that does not** is
  evidence a finding is not an artefact of the specification.

# Calibration

AUC only measures ranking. A model can rank well and still be wrong about how
likely interruption is.

```{r}
#| label: fig-calibration
#| fig-cap: "Predicted against observed frequency, pooled across resamples in deciles of predicted risk. The diagonal is perfect calibration."
#| fig-height: 4
set.seed(20260824)
calib <- map_dfr(names(sets)[names(sets) != "Structural (components)"], \(s) {
  v <- sets[[s]]
  rf <- ranger(as.formula(paste("y ~", paste(v, collapse = "+"))),
               data = tr1, probability = TRUE, num.trees = 800, seed = 1)
  tibble(`Predictor set` = s,
         p = predict(rf, te1)$predictions[, "yes"],
         obs = as.numeric(te1$y == "yes"))
})

calib |>
  mutate(bin = ntile(p, 10), .by = `Predictor set`) |>
  summarise(pred = mean(p), obs = mean(obs), n = n(),
            .by = c(`Predictor set`, bin)) |>
  ggplot(aes(pred, obs, colour = `Predictor set`)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.6) +
  geom_point(aes(size = n), alpha = 0.85) +
  scale_colour_manual(values = PAL_SET) +
  scale_size_continuous(range = c(1, 4), guide = "none") +
  scale_x_continuous(labels = label_percent()) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "Predicted probability", y = "Observed frequency", colour = NULL)
```

# Save

```{r}
#| label: save
saveRDS(list(results = results, paired = paired, summary = summ,
             importance = imp_summary, delong = dl,
             n_resamples = nrow(resamples)),
        "../data/processed/prediction_h1.rds")
```
